---
title: "STITCH and imputation-based analyses Anoles"
author: "Yann Bourgeois"
format: html
editor: visual
---

## Using STITCH to obtain imputed genotypes from BAM and a list of positions.

We can check how the imputation works by examining genotypes in the higher depth of coverage samples. Then it is a matter of looking at the confusion matrix (0/0 v 1/1)

Installation:

```{bash}

module load r

version=1.7.3
wget -O STITCH.zip https://github.com/rwdavies/STITCH/archive/refs/tags/${version}.zip ## or curl
unzip STITCH.zip && mv STITCH-${version} STITCH
cd STITCH && ./scripts/install-dependencies.sh
make install PREFIX=/shared/projects/natrix_genomics/analyses_anoles/local/STITCH

##In Makefile for IFB --library=/shared/projects/natrix_genomics/Rlibs/ . Think to add it also to the STITCH.R script

#--library=/mnt/lustre2/mres/Linum/local/STITCH
```

https://pmc.ncbi.nlm.nih.gov/articles/PMC11286293/

Then, we need to test over different values of S, K, and nGen. Read a bit https://github.com/rwdavies/STITCH/issues/59

I assume a long-term Ne of ca 50,000. Rough approximation based on the demographic trajectories. I remove shared/projects/popgen_reunion_harrier/44_S106_mdup.withrg.csorted.cleaned.aligned.bam and the high depth genomes from the bam list.\
We previously used bcftools to generate a VCF file. We can therefore start from this to generate a list of positions to examine.

Sample_AC8-8.srtd.bam has the highest depth (modal depth at 13X) among all individuals sampled in the New Orleans area, and will be used to evaluate the quality of imputation. We subsample it to a depth of roughly 2X, so sample on average one every 7 reads.

```{bash}


###The filtered vcf was generated as such:
vcftools --gzvcf calling_${SLURM_ARRAY_TASK_ID}.vcf.gz --minQ 20 --max-missing 0.6 --mac 1 --remove-indels --recode --recode-INFO-all --out filtered_calling_${SLURM_ARRAY_TASK_ID}

rm all_filtered.vcf 
for i in {1..1000};do grep -v "#" filtered_calling_${i}.recode.vcf >> all_filtered.vcf;echo $i;done

###Now we redivide by chromosome/scaffold:
#!/bin/bash
#SBATCH --mem=32GB
#SBATCH --partition=fast
#SBATCH --job-name=generate_input
#SBATCH --time=24:00:00
#SBATCH --output=generate_input.%J.out
#SBATCH --error=generate_input.%J.err
module load vcflib/1.0.3 
module load vcftools
module load r
module load bcftools
module load tabix

for chr in $(bcftools query -f '%CHROM\n' all_filtered.vcf.gz | sort -u); do
bcftools view -r "$chr" all_filtered.vcf.gz -Oz -o "${chr}.vcf.gz"
tabix -p vcf "${chr}.vcf.gz"
zcat ${chr}.vcf.gz | cut -f1,2,4,5 | grep -v "#" | grep -v "," > pos_${chr}.txt
bcftools query -s Sample_AC8-8.srtd.bam -f '%CHROM\t%POS[\t%GT]\n' ${chr}.vcf.gz  > genotypes_${chr}.txt

###The following masks triallelic sites.
awk 'BEGIN {OFS="\t"} {
    for (i=3; i<=NF; i++) {
        if ($i == "0/0" || $i == "0|0") $i = 0;
        else if ($i == "0/1" || $i == "1/0" || $i == "0|1" || $i == "1|0") $i = 1;
        else if ($i == "1/1" || $i == "1|1") $i = 2;
        else $i = "NA";  # Missing genotypes or uncalled sites
    }
    print
}' genotypes_${chr}.txt > genotypes_highdepth_numeric_${chr}.txt

awk 'NR==FNR {key[$1,$2]; next} ($1,$2) in key' pos_${chr}.txt genotypes_highdepth_numeric_${chr}.txt > matched_rows_${chr}.txt

# header Sample_AC8-8.srtd.bam

cut -f3 matched_rows_${chr}.txt | cat header - > genotypes_highdepth_numeric_${chr}.txt
done






for K in 4 8 16 24
do
for S in 1 2 5 10
do
mkdir ./imputation_${K}_${S}_$((100000 / K))
done
done



####Generate a subsample BAM:
gatk DownsampleSam \
  -I Sample_AC8-8.srtd.bam \
  -O /dev/stdout \
  -P 0.15 \
  --ACCURACY 0.0001 \
| \
gatk AddOrReplaceReadGroups \
  -I /dev/stdin \
  -O /dev/stdout \
  -RGID Sample_AC8-8_sub \
  -RGPL illumina \
  -RGSM Sample_AC8-8_sub | \
gatk SortSam \
  -I /dev/stdin \
  -O Sample_Ac8-8_sub.bam \
  -SO coordinate \
  --CREATE_INDEX true


gatk AddOrReplaceReadGroups \
  -I Sample_AC8-13.srtd.bam \
  -O Sample_Ac8-13.RG.bam \
  -RGID Sample_AC8-13 \
  -RGPL illumina \
  -RGSM Sample_AC8-13 
  
  
  
rm -f config_commands_imputation.txt
for K in 4 8 16 24
do
for S in 1 2 5 10
do
echo /shared/projects/natrix_genomics/STITCH/STITCH.R --chr=chr6 --bamlist=list_bam_imputation.txt --posfile=pos_chr6.txt --genfile=genotypes_highdepth_numeric_chr6.txt --method=diploid --outputdir=./imputation_${K}_${S}_$((100000 / K)) --K=${K} --nGen=$((100000 / K)) --S ${S} --nCores=6  --regionStart=1 --regionEnd=5000000 --buffer=10000 >> config_commands_imputation.txt
done
done

nl -w2 config_commands_imputation.txt > config_commands_imputation.txt2;mv config_commands_imputation.txt2 config_commands_imputation.txt


ls *bam | grep -v AC | grep -v TX > list_bam_imputation.txt


#!/bin/bash
#SBATCH --mem=64GB
#SBATCH --partition=fast
#SBATCH --job-name=impute
#SBATCH --time=24:00:00
#SBATCH --array=1-16   # 192 jobs
#SBATCH --cpus-per-task=6
#SBATCH --output=impute.%J.out
#SBATCH --error=impute.%J.err
module load r

config=config_commands_imputation.txt
CMD=$(awk -F'\t' -v ArrayTaskID=$SLURM_ARRAY_TASK_ID '$1==ArrayTaskID {print $2}' $config)

eval ${CMD}


```

-   **chr** What chromosome to run. Should match BAM header (the SM: field)

-   **posfile** Where to find file with positions to run. File is tab separated with no header, one row per SNP, with col 1 = chromosome, col 2 = physical position (sorted from smallest to largest), col 3 = reference base, col 4 = alternate base. Bases are capitalized. STITCH only handles bi-allelic SNPs

-   **K** Integer, how many founder / mosaic haplotypes to use

-   **nGen** Estimated number of generations since founding. In uncertain, estimate using 4 \* Ne / K, where Ne is the effective population size

-   **outputdir** What output directory to use / where output files go

-   **bamlist** (one of bamlist or cramlist required) Path to file with BAM file locations. File is one row per entry, path to BAM files. BAM index files should exist in same directory as for each BAM, suffixed either .bam.bai or .bai

-   **cramlist** (one of bamlist or cramlist required) Same as bamlist, but path is to CRAM locations. If used, requires reference variable to be set with path to fasta file

-   **S** controls the number of sets of ancestral haplotypes used. For instances where imputation confirms well to the model, S should have limited effect. For instances where imputation results are more modest, for example wild populations, increasing S should improve overall accuracy, with an approximately linear effect on run time.

-   **genfile** Path to gen file with high coverage results. Empty for no genfile. File has a header row with a name for each sample, matching what is found in the bam file. Each subject is then a tab seperated column, with 0 = hom ref, 1 = het, 2 = hom alt and NA indicating missing genotype, with rows corresponding to rows of the posfile. Note therefore this file has one more row than posfile which has no header

We can examine the Pearson's r\^2 at the end of the run. We can also compare the genotypes ourselves. For example, this run seems to have behaved well and is fast (K=8, S=10, nGen=12500), with a quite stable r2 at 0.83-0.84 over the course of the run.

So we look at the generated VCF. We can flag the positions with reasonable allele frequencies. First we go back to the original VCF, then we compute missingness and allele frequency in vcftools, we produce the genotype matrix for the individual of interest in the imputed VCF, so we can then compare it with the actual data.

```{bash}

module load vcftools
module load bcftools

##vcftools --gzvcf chr6.vcf.gz --counts --out chr6_counts_prefilter


bcftools query -s new_sample -f '%CHROM\t%POS\t%EAF\t%INFO_SCORE[\t%DS]\n' imputation_8_10_12500/stitch.chr6.1.5000000.vcf.gz > genotypes_imputed.txt
##We extract the estimated allele frequency, the INFOSCORE and the individual DOSAGE

```

We now have genotypes_imputed.txt, genotypes_highdepth_numeric.txt, and chr6_counts_prefilter.frq.count

We import all that in R:

```{r}
library(data.table)
library(tidyr)
library(dplyr)
library(ggplot2)
library(caret)  # For confusionMatrix()
library(ggplot2)
library(reshape2)

real=as.data.frame(fread("Imputation/genotypes_highdepth_numeric_chr6.txt"))
imputed=as.data.frame(fread("Imputation/genotypes_imputed.txt"))

colnames(imputed)=c("CHR","POS","EAF","INFO_SCORE","DOSAGE")
df=cbind(imputed,head(real,n=nrow(imputed)))
df=na.omit(df)


df$true_genotype <- as.factor(df$new_sample)

# Plot
n_colors <- length(unique(df$true_genotype))
palette <- scales::hue_pal()(n_colors)  # Or use RColorBrewer::brewer.pal(n, "Set3")

ggplot(df, aes(x = DOSAGE, color = true_genotype, fill = true_genotype, group = new_sample)) +
  geom_density(urban_PC1 = 0.4) +
  theme_minimal() +
  labs(x = "Dosage", y = "Density", title = "Dosage Distributions by Sample") +
  theme(legend.position = "right")


##Look at INFO score depending on dosage

ggplot(df, aes(x = INFO_SCORE, y = DOSAGE, color = as.factor(true_genotype))) +
  geom_point(urban_PC1 = 0.6, size = 2) +
  theme_minimal() +
  labs(
    x = "INFO Score",
    y = "Dosage",
    color = "True Genotype",
    title = "Dosage vs INFO Score Colored by True Genotype"
  ) +
  scale_color_brewer(palette = "Set1")  



df$geno_imputed=ifelse(df$DOSAGE<0.5,0,1)
df$geno_imputed=ifelse(df$DOSAGE>=1.5,2,df$geno_imputed)

table(df$geno_imputed,df$true_genotype)

# Function to compute confusion matrix as a data frame
compute_conf_matrix_df <- function(dosage_col, genotype_col, individual) {
  predicted <- df[[dosage_col]]  # Convert dosage to 0/1
  actual <- df[[genotype_col]]  # True genotype
  
  # Compute confusion matrix
  cm <- table(Predicted = predicted, Actual = actual)
  
  # Convert to data frame for plotting
  cm_df <- as.data.frame(cm)
  cm_df$Individual <- individual  # Add individual label
  
  return(cm_df)
}

# Compute matrices for each individual
cm <- compute_conf_matrix_df("geno_imputed", "true_genotype", "AC8_8")

# Plot the confusion matrix
ggplot(cm, aes(x = Actual, y = Predicted, fill = 100*Freq/108334)) +
  geom_tile() + 
  geom_text(aes(label = round(100*Freq/108334,2)), color = "white", size = 6) +  # Add counts
  scale_fill_gradient(low = "blue", high = "red") +  # Color scale
  labs(title = "Confusion Matrix for Genotypic Dosage Prediction",
       x = "Actual Genotype",
       y = "Predicted Genotype") +
  theme_minimal() + 



###Now, what are the results if we filter on missing data and MAF?
df_filter=subset(df,df$EAF >0.049 & df$EAF <0.951)

ggplot(df_filter, aes(x = DOSAGE, color = true_genotype, fill = true_genotype, group = new_sample)) +
  geom_density(urban_PC1 = 0.4) +
  theme_minimal() +
  labs(x = "Dosage", y = "Density", title = "Dosage Distributions by Sample") +
  theme(legend.position = "right")

# Function to compute confusion matrix as a data frame
compute_conf_matrix_df <- function(dosage_col, genotype_col, individual) {
  predicted <- df_filter[[dosage_col]]  # Convert dosage to 0/1
  actual <- df_filter[[genotype_col]]  # True genotype
  
  # Compute confusion matrix
  cm <- table(Predicted = predicted, Actual = actual)
  
  # Convert to data frame for plotting
  cm_df <- as.data.frame(cm)
  cm_df$Individual <- individual  # Add individual label
  
  return(cm_df)
}


# Compute matrices for each individual
cm <- compute_conf_matrix_df("geno_imputed", "true_genotype", "AC8_8")

# Plot the confusion matrix
ggplot(cm, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() + 
  geom_text(aes(label = Freq), color = "white", size = 6) +  # Add counts
  scale_fill_gradient(low = "blue", high = "red") +  # Color scale
  labs(title = "Confusion Matrix for Genotypic Dosage Prediction",
       x = "Actual Genotype",
       y = "Predicted Genotype") +
  theme_minimal()




```

Imputation is clearly good enough! We will take the imputed output and use a MAF cut-off of 0.05 to filter out rare variants.

We now generate the imputed VCF for all SNPs. imputation_8_10_12500, seems to provide the best results, and is fast.

```{bash}

rm -f config_commands_imputation_chr.txt

output_script="config_commands_imputation_chr.txt"
> "$output_script"  # clear or create the file

window_size=5000000
buffer=10000

for file in pos_chr*.txt; do
    chr=$(echo "$file" | sed 's/pos_\(.*\)\.txt/\1/')
    max_pos=$(tail -n 1 "$file" | awk '{print $2}')

    start=1
    while [ "$start" -lt "$max_pos" ]; do
        end=$((start + window_size - 1))
        if [ "$end" -gt "$max_pos" ]; then
            end="$max_pos"
        fi

        echo "/shared/projects/natrix_genomics/STITCH/STITCH.R --chr=${chr} --bamlist=list_bam_imputation.txt --posfile=pos_${chr}.txt --method=diploid --outputdir=./imputation_8_10_12500_${chr} --K=8 --nGen=12500 --S 10 --nCores=3 --regionStart=${start} --regionEnd=${end} --buffer=${buffer}" >> "$output_script"

        start=$((end + 1))
    done
done



nl -w2 config_commands_imputation_chr.txt > config_commands_imputation_chr.txt2;mv config_commands_imputation_chr.txt2 config_commands_imputation_chr.txt


for file in pos_chr*.txt; do
chr=$(echo "$file" | sed 's/pos_\(.*\)\.txt/\1/')
mkdir ./imputation_8_10_12500_${chr} 
done




#!/bin/bash
#SBATCH --mem=32GB
#SBATCH --partition=fast
#SBATCH --job-name=impute
#SBATCH --time=24:00:00
#SBATCH --array=1-586   # 192 jobs
#SBATCH --cpus-per-task=3
#SBATCH --output=impute.%J.out
#SBATCH --error=impute.%J.err
module load vcflib/1.0.3 
module load vcftools
module load r
module load bcftools
module load tabix

config=config_commands_imputation_chr.txt
CMD1=$(awk -F'\t' -v ArrayTaskID=$SLURM_ARRAY_TASK_ID '$1==ArrayTaskID {print $2}' $config)

eval ${CMD1}

```

And then we can replace the genotypes depending on dosage.

```{bash}
#!/bin/bash
#SBATCH --mem=64GB
#SBATCH --partition=fast
#SBATCH --job-name=final_files_imputed
#SBATCH --time=24:00:00
#SBATCH --output=final_files_imputed.%J.out
#SBATCH --error=final_files_imputed.%J.err

module load bcftools
module load vcftools
module load tabix
module load gatk4

for file in imputation_8_10_12500_*/*vcf.gz
do
basename="${file##*/}"           # Extracts "stitch.CP027624.1.vcf.gz"
name="${basename#stitch.}"       # Removes "stitch."
name="${name%.vcf.gz}"           # Removes ".vcf.gz"
echo $name
tabix -f -p vcf ${file}
bcftools +setGT ${file} -- -t "./."  -n c:'0/1' | bcftools +setGT -- -t q -i 'FMT/DS<0.5' -n c:'0/0' | bcftools +setGT -- -t q -i 'FMT/DS>=1.5' -n c:'1/1' > imputed_${name}.vcf
bcftools query -s new_sample -f '%CHROM\t%POS[\t%DS]\n' ${file} > genotypes_imputed_${name}.txt
done
cat genotypes_imputed_*.txt > genotypes_imputed_all.txt
ls imputed_*.vcf > input_variant_files.list 


for f in imputed_*.vcf; do
  bcftools view --header-only "$f" | grep -v '^##contig' > tmp_header.txt
  bcftools view "$f" | grep -v '^#' >> tmp_variants.txt
  cat tmp_header.txt tmp_variants.txt | bgzip > cleaned_${f}
  tabix -p vcf cleaned_${f}
done

bcftools concat -O z -o All_imputed.vcf.gz cleaned_*.vcf.gz
```

Now we can look at imputation quality for all sites in the genome:

```{bash}
module load bcftools
module load vcftools

rm -f genotypes_highdepth_numeric_all.txt
for chr in {1..15}
do
vcftools --vcf  Lu${chr}_no_indels.recode.vcf --max-alleles 2 --mac 1 --counts --out Lu${chr}_counts
bcftools query -s HighDepth_IOW2_8,HighDepth_5_1,HighDepth_12_6 -f '%CHROM\t%POS[\t%GT]\n' Lu${chr}_no_indels.recode.vcf > genotypes_true_${chr}.txt

###The following masks heterozygous sites.
awk 'BEGIN {OFS="\t"} {
    for (i=3; i<=NF; i++) {
        if ($i == "0/0" || $i == "0|0") $i = 0;
        else if ($i == "0/1" || $i == "1/0" || $i == "0|1" || $i == "1|0") $i = "NA";
        else if ($i == "1/1" || $i == "1|1") $i = 2;
        else $i = "NA";  # Missing genotypes or uncalled sites
    }
    print
}' genotypes_true_${chr}.txt >> genotypes_highdepth_numeric_all.txt
done

rm -f All_counts_prefilter.frq.count
for file in Lu*_counts.frq.count
do 
sed -e "s/:/\t/g" ${file} | grep -v CHROM >> All_counts_prefilter.frq.count
done

```

Now for the final check of imputation quality:

```{r}

library(data.table)
library(tidyr)
library(dplyr)
library(ggplot2)
library(caret)  # For confusionMatrix()
library(ggplot2)
library(reshape2)

real=as.data.frame(fread("genotypes_highdepth_numeric_all.txt"))
imputed=as.data.frame(fread("genotypes_imputed_all.txt"))
metadata=as.data.frame(fread("All_counts_prefilter.frq.count"))
colnames(metadata)=c("CHR","POS","N_ALLELES","N_CHR","REF","REF_COUNT","ALT","ALT_COUNT")
colnames(real)=c("CHR","POS","L65","L30","L18")

colnames(imputed)=c("CHR","POS","L65_DOSAGE","L30_DOSAGE","L18_DOSAGE")
imputed=merge(imputed,real,by=c("CHR","POS"))
df=merge(metadata,imputed,by=c("CHR","POS"))
boxplot(df$L65_DOSAGE~df$L65)

df$L65_imputed=ifelse(df$L65_DOSAGE>1,2,0)
df$L30_imputed=ifelse(df$L30_DOSAGE>1,2,0)
df$L18_imputed=ifelse(df$L18_DOSAGE>1,2,0)

df_original=df


# Function to compute confusion matrix as a data frame
compute_conf_matrix_df <- function(dosage_col, genotype_col, individual) {
  predicted <- df[[dosage_col]]  # Convert dosage to 0/1
  actual <- df[[genotype_col]]  # True genotype
  
  # Compute confusion matrix
  cm <- table(Predicted = predicted, Actual = actual)
  
  # Convert to data frame for plotting
  cm_df <- as.data.frame(cm)
  cm_df$Individual <- individual  # Add individual label
  
  return(cm_df)
}

# Compute matrices for each individual
cm_L65 <- compute_conf_matrix_df("L65_imputed", "L65", "L65")
cm_L30 <- compute_conf_matrix_df("L30_imputed", "L30", "L30")
cm_L18 <- compute_conf_matrix_df("L18_imputed", "L18", "L18")

# Combine all data frames
cm_all <- rbind(cm_L65, cm_L30, cm_L18)

# Plot the confusion matrices
ggplot(cm_all, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() + 
  geom_text(aes(label = Freq), color = "white", size = 6) +  # Add counts
  facet_wrap(~Individual) +  # One panel per individual
  scale_fill_gradient(low = "blue", high = "red") +  # Color scale
  labs(title = "Confusion Matrix for Genotypic Dosage Prediction",
       x = "Actual Genotype",
       y = "Predicted Genotype") +
  theme_minimal()



###Now, what are the results if we filter on missing data and MAF?
df=df_original
df_filter=subset(df,df$N_CHR>77 & df$ALT_COUNT/df$N_CHR >0.049 & df$ALT_COUNT/df$N_CHR <0.951)
df=df_filter
##At least 40% covered in the preimputed dataset

cm_L65 <- compute_conf_matrix_df("L65_imputed", "L65", "L65")
cm_L30 <- compute_conf_matrix_df("L30_imputed", "L30", "L30")
cm_L18 <- compute_conf_matrix_df("L18_imputed", "L18", "L18")

# Combine all data frames
cm_all <- rbind(cm_L65, cm_L30, cm_L18)

# Plot the confusion matrices
ggplot(cm_all, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() + 
  geom_text(aes(label = Freq), color = "white", size = 6) +  # Add counts
  facet_wrap(~Individual) +  # One panel per individual
  scale_fill_gradient(low = "blue", high = "red") +  # Color scale
  labs(title = "Confusion Matrix for Genotypic Dosage Prediction",
       x = "Actual Genotype",
       y = "Predicted Genotype") +
  theme_minimal()


```

## Quick test of Fst differentiation near a differentiation peak

```{bash}
vcftools --gzvcf All_imputed.vcf.gz --chr chr4 --from-bp 121500000 --to-bp 123500000 --weir-fst-pop urban_list_vcf.txt --weir-fst-pop nonurban_list_vcf.txt --fst-window-size 50000 --fst-window-step 10000 --out test_fst_chr4
```

The peak is where it is supposed to be yaaay!

## Preparing files for a FEEMS analysis

We need a file with the local grid, a set of sample coordinates, and the envelope around the study area.

First we prepare the input files with plink:

```{bash}

plink --vcf All_imputed.vcf.gz --maf 0.05 --geno 0 --make-bed --allow-extra-chr --double-id  --recode --out filtered

plink --bfile filtered --allow-extra-chr --double-id  --recodeA --out filtered

##thin every 50kb like for FEEMS
plink --bfile filtered --bp-space 50000 --allow-extra-chr --make-bed --double-id --out filtered_small

plink --bfile filtered --bp-space 50000 --allow-extra-chr --double-id  --recodeA --out filtered_small
plink --bfile filtered --bp-space 50000 --allow-extra-chr --double-id --recode --tab --out filtered_small

```

```{r}


library(dggridR)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)

library(dggridR)

filename <- dgearthgrid(dgconstruct(res=10, projection = "ISEA", aperture = 4, topology = "TRIANGLE"),savegrid = "world_triangle_res10.shp")

coord=read.table("FEEMS.coord")
outer=read.table("FEEMS.outer")
plot(outer$V1,outer$V2,col="red",pch=16)
points(coord$V1,coord$V2,col="green",pch=16)




```

```{r}
library(sf)
library(dggridR)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)

# 1. Get Louisiana + Mississippi polygons
usa <- ne_states(country = "United States of America", returnclass = "sf")
la <- usa %>% filter(name %in% c("Louisiana", "Mississippi"))

# 2. Create DGGS grid
dggs <- dgconstruct(
  res = 7,  # increase for finer resolution
  topology = "HEXAGON",
  projection = "ISEA",
  precision = 7
)

# 3. Get centroids covering bounding box
bbox <- st_bbox(la)

# Generate grid cell centers over region
centers <- dgseqnum(dggs, bbox["ymin"], bbox["xmin"], bbox["ymax"], bbox["xmax"])
cell_centers <- dgcellstolatlong(dggs, centers$seqnum)

# 4. Convert to polygons (grid cells)
grid <- dgcellstogrid(dggs, centers$seqnum)
grid_sf <- st_as_sf(grid)

# 5. Clip to Louisiana and Mississippi
grid_clipped <- st_intersection(grid_sf, la)

# 6. Plot
library(tmap)
tmap_mode("view")
tm_shape(grid_clipped) +
  tm_borders(col = "black") +
  tm_fill(col = "lightblue", urban_PC1 = 0.5) +
  tm_shape(la) +
  tm_borders(col = "red", lwd = 2)

```

## Running a RDA to better quantify the local environment

To quantify urbanization across our sites, we included 25 landscape variables to describe climatic and structural site variation. We included all 19 BIOCLIM ([76](https://www.pnas.org/doi/10.1073/pnas.2216789120#core-r76)) climate layers (v2.1) and light at night global radiance (light at night: NOAA Global Radiance Calibrated Nighttime Lights F16_20100111-20110731_rad_v4 GeoTIFF, <https://ngdc.noaa.gov/eog/dmsp/download_radcal.html>). Because both of these datasets are at a larger resolution (\~1 km^2^) and summarize variables that have diffuse effects across the landscape (e.g., effects of light at night are not constrained by site boundaries), we extracted site-level averages from each sampled area plus a 1-km buffer around the perimeter. We also included two higher-resolution (30 m^2^) land-cover layers: impervious surface cover ([77](https://www.pnas.org/doi/10.1073/pnas.2216789120#core-r77)) and canopy cover ([78](https://www.pnas.org/doi/10.1073/pnas.2216789120#core-r78)), from which we extracted site-level averages within the boundaries of each sampled site.

I do not characterize the habitat in terms of numbers of perchs like in Winchell's paper given my lack of expertise and lack of time.

How can we extract these variables? We have the file, F16_20100111-20110731_rad_v4.geotiff

We can also obtain canopy cover from 2015 and fractional impervious surface from 2015-2023 from here: https://www.mrlc.gov/data. We will use 2015 data.

```{r}

library(geodata)
library(terra)
library(adegenet)
library(vegan)
library(data.table)
library(ggplot2)
library(robust)
library(qvalue)
library(psych)
library(terra)
library(raster)
library(rworldmap)
library(dplyr)
library(caret)


gen=read.PLINK("filtered_small.raw",map.file = "filtered_small.map") ##already imputed

###Script to generate PCA to correct for pop structure later. Same script adapted with and without clones
pca_imputed <- rda(gen, scale=T) # PCA in vegan uses the rda() call without any predictors
screeplot(pca_imputed, type = "barplot", npcs=10, main="PCA Eigenvalues")
PCs <- scores(pca_imputed, choices=c(1:3), display="sites", scaling=0)
signif.axis <- anova.cca(pca_imputed, by="axis", parallel=getOption("mc.cores"))

write.table(cbind(gen@ind.names,PCs),file="PCA_imputed_small.txt",quote=F,sep="\t",row.names=F,col.names=F)





coords = read.table("FEEMS.coord",h=F)
colnames(coords)=c("lon","lat")
pca=read.table("PCA_imputed_small.txt")
colnames(pca)=c("Individual",paste("PC",seq(1:(ncol(pca)-1)),sep=""))
total=cbind(pca,coords)

colnumber=ncol(total)
for (i in 1:19){
bioclim <- terra::rast(paste("/home/yannbourgeois/Desktop/CurrentWORK/Linum/wc2.1_5m_bio/wc2.1_5m_bio_",i,".tif",sep=""))  ##Fairly low resolution, but should not matter too much at that scale
points <- terra::vect(total[,c(colnumber-1,colnumber)], geom = c("lon", "lat"), crs = "EPSG:4326")

# Extract Bioclim data, assigning nearest available value for missing points
total[,colnumber+i] <- terra::extract(bioclim, points, method = "bilinear")[,2]

colnames(total)[colnumber+i]=paste("bio",i,sep="")
}
canopy_cover=terra::rast("nlcd_tcc_CONUS_2015_v2021-4/nlcd_tcc_conus_2015_v2021-4.tif")
nightlight=terra::rast("F16_20100111-20110731_rad_v4.geotiff/F16_20100111-20110731_rad_v4.avg_vis.tif") ##the other file just lists counts to check quality of inference
perc_impervious=terra::rast("Annual_NLCD_FctImp_2015-2023_CU_C1V0/Annual_NLCD_FctImp_2015_CU_C1V0.tif")

points <- terra::vect(total[,c(colnumber-1,colnumber)], geom = c("lon", "lat"), crs = "EPSG:4326")

# Extract Bioclim data, assigning nearest available value for missing points
# Create 1000m buffer around each point (assuming CRS is in meters)
buffers <- terra::buffer(points, width = 1000)
# Extract average nightlight value within each buffer
total[,ncol(total)+1] <- terra::extract(canopy_cover, buffers, fun = mean, na.rm = TRUE)[,2]
colnames(total)[ncol(total)]="canopy_cover"
total[,ncol(total)+1] <- terra::extract(nightlight, buffers, fun = mean, na.rm = TRUE)[,2]
colnames(total)[ncol(total)]="nightlight"
total[,ncol(total)+1] <- terra::extract(perc_impervious, buffers, fun = mean, na.rm = TRUE)[,2]
colnames(total)[ncol(total)]="per_impervious"




map.polygon <- getMap(resolution = "low")

pdf(file="Supp_Figure_urban_data.pdf")
ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = canopy_cover), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()

ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = nightlight), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()


ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = per_impervious), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()

dev.off()



pca_result <- prcomp(na.omit(total[,(colnumber+1):(ncol(total)-3)]), scale. = TRUE)   ##only bioclim
summary(pca_result)  # See variance explained


pca_var <- pca_result$sdev^2
pca_var_exp <- pca_var / sum(pca_var) * 100  # Convert to percentage

# Extract PCA scores (sample coordinates)
pca_scores <- as.data.frame(pca_result$x)


# Create a scree plot
ggplot(data = data.frame(PC = 1:length(pca_var_exp), Variance = pca_var_exp), 
       aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(aes(group = 1), color = "red") +
  geom_point(color = "red") +
  labs(title = "Scree Plot of PCA",
       x = "Principal Component",
       y = "Variance Explained (%)") +
  theme_minimal()




pca_loadings <- as.data.frame(pca_result$rotation[, 1:3])
pca_loadings$Variable <- rownames(pca_loadings)

# PCA scatter plot + variable loadings
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, urban_PC1 = 0.7, color = "blue") +
  geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1*15, yend = PC2*15),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +  # Arrows for loadings
  geom_text(data = pca_loadings, aes(x = PC1*15, y = PC2*15, label = Variable),
            vjust = 1, color = "red") +
  labs(title = "PCA Biplot with Loadings",
       x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_var_exp[2], 1), "%)")) +
  theme_minimal()

ggplot(pca_scores, aes(x = PC1, y = PC3)) +
    geom_point(size = 3, urban_PC1 = 0.7, color = "blue") +
    geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1*15, yend = PC3*15),
                 arrow = arrow(length = unit(0.3, "cm")), color = "red") +  # Arrows for loadings
    geom_text(data = pca_loadings, aes(x = PC1*15, y = PC3*15, label = Variable),
              vjust = 1, color = "red") +
    labs(title = "PCA Biplot with Loadings",
         x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
         y = paste0("PC3(", round(pca_var_exp[3], 1), "%)")) +
    theme_minimal()



colnames(pca_scores)=paste("env_",colnames(pca_scores),sep="")



####And now a PCA on variables describing urban habitat:


pca_result <- prcomp(na.omit(total[,(ncol(total)-2):ncol(total)]), scale. = TRUE)   ##only bioclim
summary(pca_result)  # See variance explained


pca_var <- pca_result$sdev^2
pca_var_exp <- pca_var / sum(pca_var) * 100  # Convert to percentage

# Extract PCA scores (sample coordinates)
pca_scores2 <- as.data.frame(pca_result$x)


# Create a scree plot
ggplot(data = data.frame(PC = 1:length(pca_var_exp), Variance = pca_var_exp), 
       aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(aes(group = 1), color = "red") +
  geom_point(color = "red") +
  labs(title = "Scree Plot of PCA",
       x = "Principal Component",
       y = "Variance Explained (%)") +
  theme_minimal()




pca_loadings <- as.data.frame(pca_result$rotation[, 1:3])
pca_loadings$Variable <- rownames(pca_loadings)

# PCA scatter plot + variable loadings
ggplot(pca_scores2, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, urban_PC1 = 0.7, color = "blue") +
  geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1*15, yend = PC2*15),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +  # Arrows for loadings
  geom_text(data = pca_loadings, aes(x = PC1*15, y = PC2*15, label = Variable),
            vjust = 1, color = "red") +
  labs(title = "PCA Biplot with Loadings",
       x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_var_exp[2], 1), "%)")) +
  theme_minimal()

ggplot(pca_scores2, aes(x = PC1, y = PC3)) +
    geom_point(size = 3, urban_PC1 = 0.7, color = "blue") +
    geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1*15, yend = PC3*15),
                 arrow = arrow(length = unit(0.3, "cm")), color = "red") +  # Arrows for loadings
    geom_text(data = pca_loadings, aes(x = PC1*15, y = PC3*15, label = Variable),
              vjust = 1, color = "red") +
    labs(title = "PCA Biplot with Loadings",
         x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
         y = paste0("PC3(", round(pca_var_exp[3], 1), "%)")) +
    theme_minimal()



colnames(pca_scores2)=paste("urban_",colnames(pca_scores2),sep="")
total=cbind(total,pca_scores,pca_scores2)


####

write.table(total,file="Env_data_RDA.txt",row.names = F,quote=F,sep="\t")


# PCA scatter plot + variable loadings
ggplot(total, aes(x = PC1, y = PC2, color = paste(lon,lat))) +
  geom_point(size = 3, urban_PC1 = 0.7) +
  labs(title = "PCA genetic imputed data",
       x = paste0("PC1 (", round(100*pca_imputed$CA$eig/sum(pca_imputed$CA$eig), 1)[1], "%)"),
       y = paste0("PC2 (", round(100*pca_imputed$CA$eig/sum(pca_imputed$CA$eig), 1)[2], "%)")) +
  theme_minimal()





map.polygon <- getMap(resolution = "low")


panel3a=ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = env_PC1), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()


panel3b=ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = env_PC2), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()



panel3c=ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = env_PC3), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()


###Urban variables
panel3d=ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = urban_PC1), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()




####Genetic variation described by PCA on pruned imputed data is similar to what was obtained with ANGSD tools. 
ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = PC1), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()




ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = PC2), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()




ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = PC3), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()



```

At this stage, we can assemble Figure 1 (showing boxplot of inbreeding, PCA, environmental PCs, and FEEMS map):

```{r}
library(ggplot2)
library(ggpubr)
library(MASS)
inbreeding=read.table("Relatedness/ngsF_results.txt",h=T)
tokeep=read.table("Relatedness/low_missingness_NewOrleans.txt")
inbreeding=subset(inbreeding,inbreeding$BAM_file%in%tokeep$V1)

table=read.table("./correspondance_IDpheno_IDsequence",h=T)
coordinates=read.table("./Phenotypes.txt",h=T)
table=merge(table,coordinates,by="animal")

inbreeding=merge(table,inbreeding,by="BAM_file",all=T)
inbreeding$Season=ifelse(is.na(inbreeding$Season)==TRUE,'Countryside',inbreeding$Season)
inbreeding$long_term_status=ifelse(is.na(inbreeding$long_term_status)==T,'not assessed',inbreeding$long_term_status)


desired_order <- c('S10', 'F10','S11','S12','S13', 'S14', 'F14','Countryside')

# Reorder the Season column
inbreeding$Season <- factor(inbreeding$Season, levels = desired_order)

inbreeding_plot=ggplot(inbreeding, aes(x = Season, y = F)) +
  geom_boxplot() +
  labs(title = "",
       x = "Season",
       y = "F") +
  theme_minimal()

###Obtain list of inbreeding coefficients for individuals to be used in FST/demographic inference.
colnames(unrelated_IDs)=c("BAM_file","Depth")
inbreeding_F=merge(unrelated_IDs,inbreeding,by="BAM_file",sort=F)
inbreeding_F$BAM_file==unrelated_IDs$BAM_file


####ANOVA tests

#find optimal lambda for Box-Cox transformation 
inbreeding$urban=ifelse(inbreeding$Season=="Countryside","non-urban","urban")
bc <- boxcox(inbreeding$F~inbreeding$urban) 
#find optimal lambda for Box-Cox transformation 
(lambda <- bc$x[which.max(bc$y)])
model <- aov(((inbreeding$F^lambda-1)/lambda) ~ inbreeding$urban + inbreeding$long_term_status)
#model <- aov(inbreeding$F ~ inbreeding$urban + inbreeding$long_term_status)

plot(model)
summary(model)

###final model for paper
model <- aov(((inbreeding$F^lambda-1)/lambda) ~ inbreeding$urban + inbreeding$Season)

plot(model)
summary(model)

###Produce FIgure 1


pop<-read.table("./Structure_pcangsd/individual_info.txt",h=T)[1:71,]
C <- as.matrix(read.table("./Structure_pcangsd/PCA_NO_SNPs.cov"))
e <- eigen(C)

#plot(e$vectors[,1:2],col=pop[,3],xlab="PC1, 4.9%",ylab="PC2, 2.23%", main="individual allele frequency",pch=16, cex=2)
# Create a data frame for ggplot

df <- data.frame(PC1 = e$vectors[, 1], 
                 PC2 = e$vectors[, 2], 
                 Population = factor(pop[, 3]))  # Convert population to a factor for coloring
df$Population=ifelse(df$Population=="darkgrey","urban","non-urban")
# Plot with ggplot
PCA_plot=ggplot(df, aes(x = PC1, y = PC2, color = Population, shape = Population)) +
  geom_point(size = 3) +  # Use size = 2 and shape = 16 (solid circle) to match original plot
  labs(x = "PC1, 4.9%", y = "PC2, 2.23%", title = "") +
  theme_minimal() + scale_color_manual(values =  c("#619CFF","#C77CFF"))
 
      

library(magick)
library(grid)
library(gridExtra)
library(patchwork)
# Read the first page of the PDF as an image (you can change the number if the summary is on another page)
summary_image <- image_read_pdf("./FEEMS_plot.pdf", pages = 1)  # Density for good resolution

FEEMS_plot <- rasterGrob(as.raster(summary_image), interpolate = TRUE)


           
# A and B stacked
PanelsAB <- inbreeding_plot / PCA_plot

# Panel C: 2x2 grid
PanelC <- (panel3a | panel3b) / (panel3c | panel3d)
#PanelC= PanelC + plot_layout(guides = "collect")
# Panel D: FEEMS
PanelD <- FEEMS_plot

# Stack C and D, with relative heights
PanelsCD <- PanelC / PanelD 

# Combine A/B with C/D and add proper tags
final_plot <- (PanelsAB | PanelsCD) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 24, face = "bold"))

# Save to file
pdf("New_Figure_1.pdf", width = 15, height = 12)
print(final_plot)
dev.off()


PanelsAB <- inbreeding_plot / PCA_plot + plot_annotation(tag_levels = 'A')
PanelCD <- ((panel3a | panel3b) / (panel3c | panel3d)) / FEEMS_plot
PanelD <- FEEMS_plot

pdf(file="New_Figure_1.pdf",width=12,height=9)
final_plot <- PanelsAB | PanelCD + plot_annotation(tag_levels = c('A', 'B', 'C','D'))
final_plot
dev.off()




```

```{r}
library(data.table)

library(ggplot2)
library(dplyr)
library(vegan)
library(adegenet)
library(qvalue)
library(robust)
#library(robust,lib.loc = "./")
gen=read.PLINK("filtered_small.raw",map.file = "filtered_small.map") ##already imputed



total=read.table("Env_data_RDA.txt",h=T)

total=subset(total,total$Individual  %in% gen@ind.names)
total <- total %>%
  mutate(Individual = factor(Individual, levels = gen@ind.names)) %>%
  arrange(Individual)

total[,26:50] <- scale(total[,26:50], center=TRUE, scale=TRUE) # center=TRUE, scale=TRUE are the defaults for scale() scale for env_PC1 to env_PC3.


###Procedure for variable selection
## Null model
RDA0 <- rda(gen ~ 1 + Condition(PC1 + PC2 + PC3) ,total, scale=F) 

## Full model
RDAfull <- rda(gen ~ urban_PC1 + env_PC1 + env_PC2 + env_PC3 + Condition(PC1 + PC2 + PC3), total, scale=F )
#RDAfull <- rda(gen ~ canopy_cover + nightlight + per_impervious + env_PC1 + env_PC2 + env_PC3 + Condition(PC1 + PC2 + PC3), total, scale=F )

#To conduct the selection procedure we used the ordiR2step function of the package vegan and the following stopping criteria: variable significance of p < 0.01 using 1000 permutations, and the adjusted R2 of the global model. We run this on a small thinned dataset to get an idea since the whole dataset is huge

## Stepwise procedure with ordiR2step function
mod <- ordiR2step(RDA0, RDAfull, Pin = 0.01, R2permutations = 1000, R2scope = T)
mod$anova
###best model is the complete one, nice. about 5.56% of variance explained by environment, 9.26% by genetics
##                    R2.adj Df    AIC      F Pr(>F)   
## + urban_PC1     0.0098816  1 961.42 2.0795  0.002 **
## + env_PC1       0.0154106  1 961.73 1.6016  0.002 **
## + env_PC3       0.0183942  1 962.33 1.3225  0.002 **
## + env_PC2       0.0211519  1 962.93 1.2959  0.002 **
## <All variables> 0.0211519                           
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
complete=rda(formula = gen ~ urban_PC1 + env_PC1 + env_PC2 + env_PC3 + Condition(PC1 + PC2 + PC3), total, scale=F)
#complete=rda(formula = gen ~ urban_PC1 + env_PC1 + env_PC2 + env_PC3 + Condition(PC1 ), total, scale=F)
plot(complete, scaling=3) 
plot(complete, choices = c(1, 3),scaling=3)
vif.cca(complete)
screeplot(complete)
#signif.axis <- anova.cca(complete, by="axis", parallel=getOption("mc.cores"))
#signif.axis   ###RDA1 and RDA2 are significant.


load.rda <- scores(complete, choices=c(1:4), display="species")  # Species scores for the first three constrained axes
hist(load.rda[,1], main="Loadings on RDA1") ##Everything looks decently normal.

hist(load.rda[,2], main="Loadings on RDA2")##Everything looks decently normal.
hist(load.rda[,3], main="Loadings on RDA3") ##Everything looks decently normal.
hist(load.rda[,4], main="Loadings on RDA4") ##Everything looks decently normal.


zscore <- function(x){
  score <- (x-mean(x))/ sd(x)     # find loadings +/-z sd from mean loading     
}


positions=as.data.frame(fread("filtered_small.map"))
Z=as.data.frame(cbind(positions[,c(1,4)],zscore(load.rda[,1])))
colnames(Z)=c("CHR","POSITION","Zscore")
Z=subset(Z,Z$CHR<7)
ggplot(Z, aes(x = POSITION, y = Zscore, color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of Partial RDA results",
       x = "Genomic Position",
       y = "Z-score for RDA1 (urban-country axis)") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity



chi_sq_values <- Z$Zscore^2
lambda <- median(chi_sq_values) / qchisq(0.5, df = 1)

Z$P_val =  2 * (1 - pnorm(abs(Z$Zscore)))/max(1,lambda)


test=subset(Z,Z$CHR==2)
plot(test$POSITION,test$Zscore,xlim=c(23.5e6,25.5e6))
plot(test$POSITION,test$Zscore,xlim=c(20.5e6,30.5e6))
abline(h=3)
abline(h=3.2)

test=subset(RDA,RDA$CHR==2)
plot(test$POSITION,-log10(test$P_val),xlim=c(23.5e6,25.5e6))
abline(v=24750000)

test=subset(RDA,RDA$CHR==4)
plot(test$POSITION,-log10(test$P_val),xlim=c(121.5e6,123.5e6))
abline(v=122750000)
```

```{r}

library(data.table)
library(ggplot2)
RDA=as.data.frame(fread("RDApartial_4RDAs.txt.gz"))
#RDA=as.data.frame(fread("RDApartial_1PC.txt.gz"))
#RDA=as.data.frame(fread("RDApartial_2PCs.txt.gz"))
#RDA=as.data.frame(fread("RDApartial_nocorr.txt.gz"))
                  

#RDA2=as.data.frame(cbind(gen@chromosome,gen@position,RDA2))
colnames(RDA)=c("CHR","POSITION","P_val","Q_val")

RDAplot=subset(RDA,RDA$P_val<0.05 & RDA$CHR < 7)
ggplot(RDAplot, aes(x = POSITION, y = -log10(P_val), color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of Partial RDA results",
       x = "Genomic Position",
       y = "P-value based on Mahalanobis distance") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity




###We can also fish for outliers using the scores along distinct RDA axes.
RDA2=as.data.frame(fread("RDApartial_loadings.txt.gz"))
#RDA2=as.data.frame(fread("RDApartial_loadings_2PCs.txt.gz"))

colnames(RDA2)=c("MRK","CHR","POSITION","score_RDA1","score_RDA2","score_RDA3","score_RDA4","CCA_RDA1","CCA_RDA2","CCA_RDA3","CCA_RDA4")

zscore <- function(x){
  score <- (x-mean(x))/ sd(x)     # find loadings +/-z sd from mean loading     
}


Z=cbind(RDA2[,2:3],zscore(RDA2$score_RDA1))
colnames(Z)=c("CHR","POSITION","Zscore")

forqqplot=Z[sample(1:nrow(Z),10000),]

qqplot(qchisq(ppoints(10000), df = 1), forqqplot$Zscore^2,
       main = expression("Chi-Square Q-Q Plot"),
       xlab = "Theoretical Quantiles",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red", lty = 2)




chi_sq_values <- Z$Zscore^2
lambda <- median(chi_sq_values) / qchisq(0.5, df = 1)

Z$P_val =  2 * (1 - pnorm(abs(Z$Zscore)))/max(1,lambda)

Z=subset(Z,abs(Z$Zscore)>=2 & Z$CHR<7)

forqqplot=Z[sample(1:nrow(Z),10000),]


ggplot(Z, aes(x = POSITION, y = -log10(P_val), color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of Partial RDA results",
       x = "Genomic Position",
       y = "Z-score for RDA2 (latitude)") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity




###At last, we can regenerate Mahalanobis distance-based p-values using a reduced set of axes (for example the first 2 RDA axes which are significant)


rdadapt<-function(rda,K)
{
#  loadings<-rda$CCA$v[,1:as.numeric(K)] #original code for on the fly calculation
  loadings<-rda[,1:as.numeric(K)] #original code for on the fly calculation
  resscale <- apply(loadings, 2, scale)
  resmaha <- covRob(resscale, distance = TRUE, na.action= na.omit, estim="pairwiseGK")$dist
  lambda <- median(resmaha)/qchisq(0.5,df=K)
  reschi2test <- pchisq(resmaha/lambda,K,lower.tail=FALSE)
  qval <- qvalue(reschi2test)
  q.values_rdadapt<-qval$qvalues
  return(data.frame(p.values=reschi2test, q.values=q.values_rdadapt))
}

RDA_2_RDA<-rdadapt(as.matrix(RDA2[,c(8,11)]), 2) ###Will output P-values corrected for inflation and qvalues, using the Mahala distance for the 2 RD strongly correlated with the urban PC.
RDA_2_RDA$CHR=RDA$CHR
RDA_2_RDA$POSITION=RDA$POSITION

colnames(RDA_2_RDA)=c("P_val","Q_val","CHR","POSITION")

RDAplot=subset(RDA_2_RDA,RDA_2_RDA$P_val<0.05)

ggplot(RDAplot, aes(x = POSITION, y = -log10(P_val), color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of Partial RDA results",
       x = "Genomic Position",
       y = "P-value based on Mahalanobis distance (2 RDA axes)") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity

write.table(RDA_2_RDA,file="RDApartial_2PCs_2urbanRDaxes.txt",row.names=F,quote=F,sep="\t")


```

We can also visualize the outlier loci found with the Mahalanobis distance in comparison with neutral loci using a RDA biplot:

```{r}
library(GenomicRanges)
library(data.table)
TAB_loci <- RDA2[,c(2:7)] # vegan references "species", here these are the loci


df_intervals=as.data.frame(fread("Selection_tests/candidate_regions_IWTomics.bed"))[,1:3]
df_intervals=as.data.frame(fread("Selection_tests/candidate_windows_Winchell.bed"))[,1:3]
colnames(df_intervals)=c("chr","start","end")
# Create GRanges objects
gr_intervals <- GRanges(seqnames = df_intervals$chr,
                        ranges = IRanges(start = df_intervals$start, end = df_intervals$end))
TAB_loci$CHR=ifelse(TAB_loci$CHR==1,"chr1",TAB_loci$CHR)
TAB_loci$CHR=ifelse(TAB_loci$CHR==2,"chr2",TAB_loci$CHR)
TAB_loci$CHR=ifelse(TAB_loci$CHR==3,"chr3",TAB_loci$CHR)
TAB_loci$CHR=ifelse(TAB_loci$CHR==4,"chr4",TAB_loci$CHR)
TAB_loci$CHR=ifelse(TAB_loci$CHR==5,"chr5",TAB_loci$CHR)
TAB_loci$CHR=ifelse(TAB_loci$CHR==6,"chr6",TAB_loci$CHR)

gr_positions <- GRanges(seqnames = TAB_loci$CHR,
                        ranges = IRanges(start = TAB_loci$POSITION, end = TAB_loci$POSITION))

# Find overlaps
hits <- findOverlaps(gr_positions, gr_intervals)

# Mark positions that have at least one overlap
TAB_loci$in_interval <- 1:nrow(TAB_loci) %in% queryHits(hits)
TAB_loci$type =ifelse(RDA$Q_val<0.2,"RDA outlier only","Neutral") 
TAB_loci$type =ifelse(TAB_loci$in_interval==TRUE & RDA$Q_val<0.2,"PBS outlier and RDA outlier",TAB_loci$type)
TAB_loci$type =ifelse(TAB_loci$in_interval==TRUE & RDA$Q_val>=0.2,"PBS outlier but not RDA outlier",TAB_loci$type) 

boxplot(TAB_loci$score_RDA1~TAB_loci$type)
save=TAB_loci

TAB_var <- read.table("RDA_biplot_3PCs",h=T,row.names = 1)
#as.data.frame(scores(complete, choices=c(1:4), display="bp")) # Pulls the biplot scores from the reduced dataset from before. I guess this could also be obtained from the full dataset since we export it I think when summarizing the data. Is probably in the *err or *out job output files. 

selected=subset(TAB_loci,TAB_loci$type=="PBS outlier but not RDA outlier")
neutral=subset(TAB_loci,TAB_loci$type=="Neutral")[sample(10000),]
TAB_loci=rbind(neutral,selected)
## Biplot of RDA loci and variables scores
ggplot() +
  geom_hline(yintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_vline(xintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_point(data = TAB_loci, aes(x=score_RDA1*20, y=score_RDA2*20, colour = type), size = 1.4) +
  scale_color_manual(values = c("lightblue","darkred", "#F9A242FF", "#6B4596FF")) +
  geom_segment(data = TAB_var, aes(xend=RDA1, yend=RDA2, x=0, y=0), colour="black", size=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
  geom_text(data = TAB_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(TAB_var)), size = 2.5, family = "Times") +
  xlab("RDA 1") + ylab("RDA 2") +
  facet_wrap(~"RDA space") +
  guides(color=guide_legend(title="Locus type")) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.background = element_blank(), legend.background = element_blank(), panel.grid = element_blank(), plot.background = element_blank(), legend.text=element_text(size=rel(.8)), strip.text = element_text(size=11))




ggplot() +
  geom_hline(yintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_vline(xintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_point(data = TAB_loci, aes(x=score_RDA3*20, y=score_RDA4*20, colour = type), size = 1.4) +
  scale_color_manual(values = c("lightblue","darkred", "#F9A242FF", "#6B4596FF")) +
  geom_segment(data = TAB_var, aes(xend=RDA3, yend=RDA4, x=0, y=0), colour="black", size=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
  geom_text(data = TAB_var, aes(x=1.1*RDA3, y=1.1*RDA4, label = row.names(TAB_var)), size = 2.5, family = "Times") +
  xlab("RDA 3") + ylab("RDA 4") +
  facet_wrap(~"RDA space") +
  guides(color=guide_legend(title="Locus type")) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.background = element_blank(), legend.background = element_blank(), panel.grid = element_blank(), plot.background = element_blank(), legend.text=element_text(size=rel(.8)), strip.text = element_text(size=11))


#write.table(subset(TAB_loci,TAB_loci$type=="Outlier"),file="RDA_outliers.txt",row.names=F,quote=F,sep="\t") ###311 variants
```

Ultimately we try models with 3 PC for structure correction and without structure correction. Remember, we only focused on the non-clonal dataset. We compute the p-value for Mahalanobis distance on the first 2 RDA axes (which are significant). Note that we could include more axes, but this may also incorporate more noise.

We can then visualize adaptive gradients in the landscape. We can do this for RDA, LFMM, and RSB outliers.

For example, for the RDA outliers obtained above, we can extract them from the plink file as follows:

```{bash}

cut -f1 RDA_outliers.txt | grep -v MRK | cut -f1 -d "_" | sort > outliers_rda2_snps.txt


cut -f1,2 list_individuals_tokeep_selection.txt | sed "s:L0:L:" > list_no_clones.txt 

##An issue with this dataset is the presence of many clonal individuals, usually from the same locality. 
plink --bfile filtered_all --keep list_no_clones.txt --extract  outliers_rda2_snps.txt --geno 0 --make-bed --allow-extra-chr --double-id  --recode --out outliers_rda2

plink --bfile outliers_rda2 --allow-extra-chr --double-id  --recodeA --out outliers_rda2
plink --bfile outliers_rda2 --allow-extra-chr --double-id --recode --tab --out outliers_rda2


```

We can have a look at how variance at these loci aligns with latitude. It is a bit redundant with using the RDA loadings (see below), but we do not constrain anything and it is a bit easier to interpret:

```{r}

gen=read.PLINK("outliers_rda2.raw",map.file = "outliers_rda2.map") ##already imputed

total=read.table("Env_data_RDA_no_clones.txt",h=T)

total=subset(total,total$Individual  %in% gen@ind.names)
total <- total %>%
  mutate(Individual = factor(Individual, levels = gen@ind.names)) %>%
  arrange(Individual)


pca_imputed <- rda(gen, scale=T) # PCA in vegan uses the rda() call without any predictors
screeplot(pca_imputed, type = "barplot", npcs=10, main="PCA Eigenvalues")
PCs <- as.data.frame(scores(pca_imputed, choices=c(1:3), display="sites", scaling=0))
PCs$Individual=row.names(PCs)
coords=total[,c(1,2,3,4,11,12)]
PCs=merge(PCs,coords,by="Individual")

map.polygon <- getMap(resolution = "low")


ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black", urban_PC1 = 1) +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = PCs, aes(x = lon, y = lat, fill = PC1.x), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-11, 55) + 
    ylim(29, 55) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()






# PCA scatter plot + variable loadings
ggplot(PCs, aes(x = PC1.x, y = PC2.x)) +
  geom_point(data = PCs,aes(x = PC1.x, y = PC2.x, fill=PC1.y),size = 5, shape = 21, color = "black") +
  labs(title = "PCA Biplot with Loadings",
       x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_var_exp[2], 1), "%)")) +
      scale_fill_gradient2(low = "darkblue", high = "red") +  

  theme_minimal()





# PCA scatter plot + variable loadings
ggplot(PCs, aes(x = PC1.x, y = PC2.x)) +
  geom_point(data = PCs,aes(x = PC1.x, y = PC2.x, fill=PC2.y),size = 5, shape = 21, color = "black") +
  labs(title = "PCA Biplot with Loadings",
       x = paste0("PC1 (", round(pca_var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_var_exp[2], 1), "%)")) +
      scale_fill_gradient2(low = "darkblue", high = "red") +  

  theme_minimal()


```

### Bonus, adaptability index in space.

Now, in R we can have fun projecting adaptability indexes in space.

Taken from Capblancq's Github:

"The scores of the environmental variables along the RDA axes can be used to calculate a genetic-based index of adaptation for each environmental pixel of the landscape. This index is estimated independently for each RDA axis of interest using the formula:\
$\sum_{i = 1}^{n}{a_ib_i}$ where *a* is the climatic variable score (loading) along the RDA axis, *b* is the standardized value for this particular variable at the focal pixel, and *i* refers to one of the *n* different variables used in the RDA model.

Here we can clearly see that the RD2 axis correlates with latitude

```{r}

library(dplyr)
library(vegan)
library(adegenet)
library(qvalue)
library(robust)
library(rnaturalearth)
#library(robust,lib.loc = "./")

#### Function to project the adaptive component turnover across the landscape
adaptive_index <- function(RDA, K, env_pres, range = NULL, method = "loadings", scale_env, center_env){
  
  # Formatting environmental rasters for projection
  var_env_proj_pres <- as.data.frame(rasterToPoints(env_pres[[row.names(RDA$CCA$biplot)]]))
  
  # Standardization of the environmental variables
  var_env_proj_RDA <- as.data.frame(scale(var_env_proj_pres[,-c(1,2)], center_env[row.names(RDA$CCA$biplot)], scale_env[row.names(RDA$CCA$biplot)]))
  
  # Predicting pixels genetic component based on RDA axes
  Proj_pres <- list()
  if(method == "loadings"){
    for(i in 1:K){
      ras_pres <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(apply(var_env_proj_RDA[,names(RDA$CCA$biplot[,i])], 1, function(x) sum( x * RDA$CCA$biplot[,i])))), crs = crs(env_pres))
      names(ras_pres) <- paste0("RDA_pres_", as.character(i))
      Proj_pres[[i]] <- ras_pres
      names(Proj_pres)[i] <- paste0("RDA", as.character(i))
    }
  }
  
  # Prediction with RDA model and linear combinations
  if(method == "predict"){ 
    pred <- predict(RDA, var_env_proj_RDA[,names(RDA$CCA$biplot[,i])], type = "lc")
    for(i in 1:K){
      ras_pres <- rasterFromXYZ(data.frame(var_env_proj_pres[,c(1,2)], Z = as.vector(pred[,i])), crs = crs(env_pres))
      names(ras_pres) <- paste0("RDA_pres_", as.character(i))
      Proj_pres[[i]] <- ras_pres
      names(Proj_pres)[i] <- paste0("RDA", as.character(i))
    }
  }
  
  # Mask with the range if supplied
  if(!is.null(range)){
    Proj_pres <- lapply(Proj_pres, function(x) mask(x, range))
  }
  
  # Returning projections for current climates for each RDA axis
  return(Proj_pres = Proj_pres)
}

gen=read.PLINK("outliers_rda2.raw",map.file = "outliers_rda2.map") ##already imputed

total=read.table("Env_data_RDA_no_clones.txt",h=T)

total=subset(total,total$Individual  %in% gen@ind.names)
total <- total %>%
  mutate(Individual = factor(Individual, levels = gen@ind.names)) %>%
  arrange(Individual)

total[,c(11,13:ncol(total))] <- scale(total[,c(11,13:ncol(total))], center=TRUE, scale=TRUE) # center=TRUE, scale=TRUE are the defaults for scale() scale

Env <- scale(total[,c(11,13:ncol(total))], center=TRUE, scale=TRUE) # center=TRUE, scale=TRUE are the defaults for scale() scale

scale_env <- attr(Env, 'scaled:scale')
center_env <- attr(Env, 'scaled:center')



##Run RDA only on outliers, can be obtained with plink easily
RDA_outliers <- rda(formula = gen ~ lat + env_PC1  + env_PC2 + env_PC3 + Condition(PC1 + PC2 + PC3), total, scale=F)

summary(RDA_outliers) ###Nice constrained variance you got there...


####Preparing a raster of environmental PCs and latitude.


coords = read.table("Localities.txt",h=T) ##Corrected coordinates for Greece.
pca=read.table("PCA_imputed_no_clones.txt")

coords$Individual <- gsub("L0([1-9])", "L\\1", coords$seq_code)
colnames(pca)=c("Individual",paste("PC",seq(1:(ncol(pca)-1)),sep=""))
total=merge(pca,coords,by="Individual")
total=subset(total,total$Individual!="L88")


for (i in 1:19){
bioclim <- terra::rast(paste("wc2.1_5m_bio/wc2.1_5m_bio_",i,".tif",sep=""))  ##Fairly low resolution, but should not matter too much at that scale
points <- terra::vect(total[,c(12,11)], geom = c("lon", "lat"), crs = "EPSG:4326")

# Extract Bioclim data, assigning nearest available value for missing points
total[,12+i] <- terra::extract(bioclim, points, method = "bilinear")[,2]

colnames(total)[12+i]=paste("bio",i,sep="")
}



pca_result <- prcomp(na.omit(total[,c(11,13:ncol(total))]), scale. = TRUE)
summary(pca_result)  # See variance explained

####We need the raster now. Then we need to project the points into the PCA of bioclim data obtained for the locations included in the analyses, and we create the 3 rasters corresponding to the 3 PC projections. Plus the latitude. Fun...

ras <-stack(list.files("./wc2.1_5m_bio/", pattern = ".tif$", full.names = T))
names(ras) <- gsub("_", "", unlist(strsplit(unlist(lapply(strsplit(list.files("./wc2.1_5m_bio", pattern = ".tif$", full.names = T), split = "./wc2.1_5m_bio/wc2.1_5m_"), function(x) x[2])), split = ".tif")))

crop_extent <- extent(-20, 60, 25, 58)
ras_cropped <- crop(ras, crop_extent)

latitude_layer <- ras_cropped[[1]]

# Replace cell values with their corresponding latitudes
latitude_layer[] <- coordinates(latitude_layer)[, 2]

# Assign a name to the new layer
names(latitude_layer) <- "lat"

ras_final <- addLayer(ras_cropped, latitude_layer)


remove.NAs.stack<-function(rast.stack){
  nom<-names(rast.stack)
  test1<-calc(rast.stack, fun=sum)
  test1[!is.na(test1)]<-1
  test2<-rast.stack*test1
  test2<-stack(test2)
  names(test2)<-nom
  return(test2)
}
ras_final <- remove.NAs.stack(ras_final)


pca_proj <- predict(ras_final, pca_result, index = 1:3)  # Project first 3 PCs

# pca_proj is a RasterStack with layers for PC1, PC2, PC3
names(pca_proj) <- c("env_PC1", "env_PC2", "env_PC3")
#writeRaster(pca_proj, filename = "pca_bioclim_projection.tif", format = "GTiff", overwrite = TRUE)
pca_proj_final <- addLayer(pca_proj, latitude_layer)


## Running the function for all the climatic pixels of lodgepole pine distribution range
res_RDA_proj_current <- adaptive_index(RDA = RDA_outliers, K = 2, env_pres = pca_proj_final, range = NULL, method = "loadings", scale_env = scale_env, center_env = center_env)



## Vectorization of the climatic rasters for ggplot
RDA_proj <- list(res_RDA_proj_current$RDA1, res_RDA_proj_current$RDA2)
RDA_proj <- lapply(RDA_proj, function(x) rasterToPoints(x))
for(i in 1:length(RDA_proj)){
  RDA_proj[[i]][,3] <- (RDA_proj[[i]][,3]-min(RDA_proj[[i]][,3]))/(max(RDA_proj[[i]][,3])-min(RDA_proj[[i]][,3]))
}

## Adaptive genetic turnover projected across lodgepole pine range for RDA1 and RDA2 indexes
TAB_RDA <- as.data.frame(do.call(rbind, RDA_proj[1:2]))
colnames(TAB_RDA)[3] <- "value"
TAB_RDA$variable <- factor(c(rep("RDA1", nrow(RDA_proj[[1]])), rep("RDA2", nrow(RDA_proj[[2]]))), levels = c("RDA1","RDA2"))
admin <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(data = TAB_RDA) + 
  geom_sf(data = admin, fill=gray(.9), size=0) +
  geom_raster(aes(x = x, y = y, fill = cut(value, breaks=seq(0, 1, length.out=10), include.lowest = T))) + 
  scale_fill_viridis_d(urban_PC1 = 0.8, direction = -1, option = "A", labels = c("Negative scores","","","","Intermediate scores","","","","Positive scores")) +
  geom_sf(data = admin, fill=NA, size=0.1) +
  coord_sf(xlim = c(-20, 60), ylim = c(25, 58), expand = FALSE) +
  xlab("Longitude") + ylab("Latitude") +
  guides(fill=guide_legend(title="Adaptive index")) +
  facet_grid(~ variable) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))


crop_extent <- extent(-20, 60, 25, 58)

```

## LFMM2

```{r}

library(data.table)

Y <- as.matrix(fread("filtered.lfmm"))  # or however you're loading it
snp_chunks <- lapply(1:50, function(start) Y[, seq(start, ncol(Y), by = 50)])

dir.create("lfmm_chunks")
for (i in 1:50) {
  fwrite(as.data.frame(snp_chunks[[i]]), file = paste0("lfmm_chunks/chunk_", i, ".lfmm"), sep = "\t")
}



args <- commandArgs(trailingOnly = TRUE)
chunk_id <- as.integer(args[1])
Y <- as.matrix(fread(paste0("lfmm_chunks/chunk_", chunk_id, ".lfmm")))

library(LEA)
library(data.table)

X <- readRDS("env_matrix.rds")  # assumed pre-scaled, saved once
K <- 3  # or parameterized per chunk

mod <- lfmm2(input = Y, env = X, K = K)

saveRDS(mod, file = paste0("results/mod_lfmm_chunk_", chunk_id, ".rds"))

library(LEA)
library(adegenet)
library(dplyr)
library(data.table)
###First convert the genlight object into a LEA object
ped2lfmm("filtered_small.ped") ##to be changed to take the entire dataset
Y=as.data.frame(fread("filtered_small.lfmm"))

gen=read.PLINK("filtered_small.raw",map.file = "filtered_small.map") ##already imputed

total=read.table("Env_data_RDA.txt",h=T)
total=subset(total,total$Individual  %in% gen@ind.names)
total <- total %>%
  mutate(Individual = factor(Individual, levels = gen@ind.names)) %>%
  arrange(Individual)

# # center and scale. Mostly to keep things consistent with previous pipeline.
X <- scale(total[,26:50], center=TRUE, scale=TRUE)
colnames(X)[which(colnames(X)=="urban_PC1")]="urban_PC1"

mod.lfmm2.K3.all <- lfmm2(input = Y, env = X[,c(which(colnames(X)=="urban_PC1"),which(colnames(X)=="canopy_cover"),which(colnames(X)=="nightlight"),which(colnames(X)=="per_impervious"),which(colnames(X)=="env_PC1"),which(colnames(X)=="env_PC2"),which(colnames(X)=="env_PC3"))], K = 3)

mod.lfmm2.K3.PC <- lfmm2(input = Y, env = X[,c(which(colnames(X)=="urban_PC1"),which(colnames(X)=="env_PC1"),which(colnames(X)=="env_PC2"),which(colnames(X)=="env_PC3"))], K = 3)


pv <- lfmm2.test(object = mod.lfmm2.K3.PC,
input = Y,
env = X[,c(which(colnames(X)=="urban_PC1"),which(colnames(X)=="env_PC1"),which(colnames(X)=="env_PC2"),which(colnames(X)=="env_PC3"))],
full = TRUE) ###Will produce a unique P-value for association with the whole set of independent environmental variables

pv_1 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="urban_PC1")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.


pv_2 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="canopy_cover")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.

pv_3 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="nightlight")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.


pv_4 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="per_impervious")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.

pv_5 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="env_PC1")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.

pv_6 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="env_PC2")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.

pv_7 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)=="env_PC3")],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.

positions=as.data.frame(fread("filtered_small.map"))[,c(1,4)]

All_results=as.data.frame(cbind(gen@loc.names,positions,
pv_1$pvalues,
pv_2$pvalues,
pv_3$pvalues,
pv_4$pvalues,
pv_5$pvalues,
pv_6$pvalues,
pv_7$pvalues,
pv$pvalues,
pv_1$zscores,
pv_2$zscores,
pv_3$zscores,
pv_4$zscores,
pv_5$zscores,
pv_6$zscores,
pv_7$zscores))
colnames(All_results)[1]="MRK"
colnames(All_results)[2]="CHR"
colnames(All_results)[3]="POSITION"
colnames(All_results)[4:10]=c("urban_PC1","canopy_cover","nightlight","per_impervious","env_PC1","env_PC2","env_PC3")
colnames(All_results)[11]="P_val_combined"
colnames(All_results)[12:18]=paste("zscore_",c("urban_PC1","canopy_cover","nightlight","per_impervious","env_PC1","env_PC2","env_PC3"),sep="")

                                                                                                                                                                                              
write.table(All_results,file="LFMM_scan_K3_small.txt",row.names = F,quote=F,sep="\t")


test=subset(All_results,All_results$CHR==2)
plot(test$POSITION,-log10(test$P_val_combined),xlim=c(23.5e6,25.5e6))
abline(v=24750000)



test=subset(All_results,All_results$CHR==4)
plot(test$POSITION,-log10(test$P_val_combined),xlim=c(121.5e6,123.5e6))
abline(v=122750000)



truc=subset(All_results,All_results$CHR<7)
ggplot(truc, aes(x = POSITION, y = -log10(urban_PC1), color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of Partial RDA results",
       x = "Genomic Position",
       y = "Z-score for RDA1 (urban-country axis)") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity


```

```{bash}
 for i in {1..50}; do sed -n '1d' LFMM_scan_K3_${i}.txt >> LFMM_scan_K3_parallel.txt;done
head -n1 LFMM_scan_K3_1.txt > header
sort -k2,2 -k3,3n  LFMM_scan_K3_parallel.txt | cat header - > LFMM_scan_K3_parallel.txt2;mv LFMM_scan_K3_parallel.txt2 LFMM_scan_K3_parallel.txt
gzip LFMM_scan_K3_parallel.txt
```

Another way to parallelize:

```{r}
library(LEA)
library(adegenet)
library(dplyr)
library(data.table)
###First convert the genlight object into a LEA object
#ped2lfmm("filtered_small.ped") ##to be changed to take the entire dataset

args <- commandArgs(trailingOnly = TRUE)
analysis_id <- as.integer(args[1])
#Y <- as.matrix(fread(paste0("lfmm_chunks/chunk_", chunk_id, ".lfmm")))

Y=as.data.frame(fread("filtered.lfmm"))
#Y=as.data.frame(fread(paste0("lfmm_chunks/chunk_", chunk_id, ".lfmm")))

gen=read.PLINK("filtered.raw",map.file = "filtered.map") ##already imputed

total=read.table("Env_data_RDA.txt",h=T)
total=subset(total,total$Individual  %in% gen@ind.names)
total <- total %>%
  mutate(Individual = factor(Individual, levels = gen@ind.names)) %>%
  arrange(Individual)

# # center and scale. Mostly to keep things consistent with previous pipeline.
X <- scale(total[,26:50], center=TRUE, scale=TRUE)

positions=as.data.frame(fread("filtered.map"))[,c(1,4)]
variables=c("urban_PC1","canopy_cover","nightlight","per_impervious","env_PC1","env_PC2","env_PC3")

###Multivariate test
if (analysis_id==8){
mod.lfmm2.K3.PC <- lfmm2(input = Y, env = X[,c(which(colnames(X)=="urban_PC1"),which(colnames(X)=="env_PC1"),which(colnames(X)=="env_PC2"),which(colnames(X)=="env_PC3"))], K = 3)


pv <- lfmm2.test(object = mod.lfmm2.K3.PC,
input = Y,
env = X[,c(which(colnames(X)=="urban_PC1"),which(colnames(X)=="env_PC1"),which(colnames(X)=="env_PC2"),which(colnames(X)=="env_PC3"))],
full = TRUE) ###Will produce a unique P-value for association with the whole set of independent environmental variables


All_results=as.data.frame(cbind(gen@loc.names,positions,pv$pvalues))
colnames(All_results)[1]="MRK"
colnames(All_results)[2]="CHR"
colnames(All_results)[3]="POSITION"
colnames(All_results)[4]="P_val_combined"
write.table(All_results,file="LFMM_scan_K3_multinomial.txt",row.names = F,quote=F,sep="\t")
} else {
mod.lfmm2.K3.all <- lfmm2(input = Y, env = X[,which(colnames(X)==variables[analysis_id])], K = 3)
###Univariate test

pv_1 <- lfmm2.test(object = mod.lfmm2.K3.all,
input = Y,
env = X[,which(colnames(X)==variables[analysis_id])],
full = FALSE) ###Will produce one P-value for association with each environmental variable, we include more than above to test.
All_results=as.data.frame(cbind(gen@loc.names,positions,pv_1$pvalues,pv_1$zscores))
colnames(All_results)[1]="MRK"
colnames(All_results)[2]="CHR"
colnames(All_results)[3]="POSITION"
colnames(All_results)[4]=variables[analysis_id]
colnames(All_results)[5]=paste("zscore_",variables[analysis_id],sep="")
write.table(All_results,file=paste("LFMM_scan_K3_",variables[analysis_id],".txt",sep=""),row.names = F,quote=F,sep="\t")
}

```

We can prepare the input file that is compatible with WZA.

```{bash}
##combine outputs
paste LFMM_scan_K3_multinomial.txt \
<(cut -f4,5 LFMM_scan_K3_canopy_cover.txt) \
<(cut -f4,5 LFMM_scan_K3_env_PC1.txt) \
<(cut -f4,5 LFMM_scan_K3_env_PC2.txt) \
<(cut -f4,5 LFMM_scan_K3_env_PC3.txt) \
<(cut -f4,5 LFMM_scan_K3_nightlight.txt) \
<(cut -f4,5 LFMM_scan_K3_per_impervious.txt) \
<(cut -f4,5 LFMM_scan_K3_urban_PC1.txt) <(awk '{print $5}' filtered.frq) | awk '{window=int($3/50000)*50000; print $0 "\t" $2 "_" window}' | cut -f2- > LFMM_scan_K3_total.txt
gzip LFMM_scan_K3_total.txt

```

```{r}
library(data.table)
library(ggplot2)





test=subset(All_results,All_results$CHR==2)
plot(test$POSITION,-log10(test$P_val_combined),xlim=c(23.5e6,25.5e6))
abline(v=24750000)



test=subset(All_results,All_results$CHR==4)
plot(test$POSITION,-log10(test$P_val_combined),xlim=c(121.5e6,123.5e6))
abline(v=122750000)



truc=subset(All_results,All_results$CHR<7 & All_results$P_val_combined <0.05)
ggplot(truc, aes(x = POSITION, y = -log10(P_val_combined), color = CHR)) +
  geom_point() + # Points on the line
  facet_wrap(~CHR) +  # Separate chromosomes
  labs(title = "Manhattan Plot of LFMM2 results",
       x = "Genomic Position",
       y = "P-value for multinomial test") +
  #scale_color_manual(values = rep(c("black", "blue"),14)) +
  theme_minimal() +  theme(legend.position = "none")  # Remove legend for clarity

```

###Script for start of Figure 5 Now, we prepare Figure 5. Let's examine different sets of candidate genes identified in other studies on urbanization or based on GO terms (list of GO terms in list file, identified through a search in AMIGO)

```{r}

#Find overlap
library(GenomicRanges)
library(IRanges)
library(data.table)
library(regioneR)
library(ggpubr)
library(biomaRt)
library(dplyr)

#ensembl <- useMart("ensembl")
#listDatasets(ensembl)
ensembl <- useMart("ensembl",host="https://may2024.archive.ensembl.org", dataset = "acarolinensis_gene_ensembl")  
library(GOfuncR)
getBM(mart=ensembl,
      attributes=c('ensembl_gene_id','external_gene_name', 'go_id', 'name_1006'),
      filters="external_gene_name",
      values="Tl")
anno <- getBM(mart=ensembl,attributes=c("ensembl_gene_id", "external_gene_name", "go_id", "chromosome_name", "start_position", "end_position", "strand", "description"))

go_annot=anno[,c(1,3)]
colnames(go_annot)=c("gene","go_id")
custom_coords=anno[,c(1,4,5,6)]
custom_coords$chromosome_name=gsub('.1','',custom_coords$chromosome_name)
colnames(custom_coords)=c("gene","chr","start","end")


###Look at FST, PBS, diversity in these candidate genes. 50kb resolution, look at shorter windows
Locomotion=read.table("Selection_tests/LimbLocomotion_GOterms_list",sep="\t")
Behavior=read.table("Selection_tests/Behavir_GOterms_list",sep="\t")
Immunity=read.table("Selection_tests/Immune_GOterms_list",sep="\t")
Heat=read.table("Selection_tests/Heat_GOterms_list",sep="\t")

list_Locomotion=subset(go_annot,go_annot$go_id %in% Locomotion$V1)
list_Behavior=subset(go_annot,go_annot$go_id %in% Behavior$V1)
list_Immunity=subset(go_annot,go_annot$go_id %in% Immunity$V1)
list_Heat=subset(go_annot,go_annot$go_id %in% Heat$V1)

list_candidates_cristatellus_extended=read.table("Selection_tests/list_all_genes_morpho_urban_cristatellus")
list_candidates_cristatellus=read.table("Selection_tests/list_core_genes_cristatellus")


custom_coords=unique(custom_coords)

custom_coords$category_immune=ifelse(custom_coords$gene%in%list_Immunity$gene,1,0)
custom_coords$category_behaviour=ifelse(custom_coords$gene%in%list_Behavior$gene,1,0)
custom_coords$category_locomotion=ifelse(custom_coords$gene%in%list_Locomotion$gene,1,0)
custom_coords$category_heat=ifelse(custom_coords$gene%in%list_Heat$gene,1,0)
custom_coords$category_extended_cristal=ifelse(custom_coords$gene%in%list_candidates_cristatellus_extended$V1,1,0)
custom_coords$category_strict_cristal=ifelse(custom_coords$gene%in%list_candidates_cristatellus$V1,1,0)



custom_coords=GRanges(custom_coords)

###Generating background and candidate windows
background=subset(custom_coords,custom_coords$category_immune+ custom_coords$category_behaviour+custom_coords$category_locomotion +custom_coords$category_extended_cristal +custom_coords$category_strict_cristal==0)
immune=subset(custom_coords,custom_coords$category_immune==1)
behav=subset(custom_coords,custom_coords$category_behaviour==1)
heat=subset(custom_coords,custom_coords$category_heat==1)
locom=subset(custom_coords,custom_coords$category_locomotion==1)
extended=subset(custom_coords,custom_coords$category_extended_cristal==1)
narrow=subset(custom_coords,custom_coords$category_strict_cristal==1)


sumstats_1=as.data.frame(fread("Selection_tests/All_results_50kb_FST_PBS_ANGSD_countryside.txt",h=T))
sumstats_2=as.data.frame(fread("Selection_tests/urban_nucdiv_50kb.txt",h=F))
sumstats_3=as.data.frame(fread("Selection_tests/countryside_nucdiv_50kb.txt",h=F))
sumstats_4=as.data.frame(fread("Selection_tests/FastPCA_results.txt",h=F))

colnames(sumstats_2)=c("index","chr","WinCenter",	"tW",	"tP"	,"tF",	"tH",	"tL",	"Tajima",	"fuf",	"fud",	"fayh",	"zeng",	"nSites")
colnames(sumstats_3)=c("index","chr","WinCenter",	"tW",	"tP"	,"tF",	"tH",	"tL",	"Tajima",	"fuf",	"fud",	"fayh",	"zeng",	"nSites")
colnames(sumstats_4)=c("chr","start","end","P")

##sed -i -e "s:chrLG::" *input_*.txt
##sed -i -e "s:chrUn_::" *input_*.txt
##sed -i -e "s:chr::" *input_*.txt


sumstats_1$start=sumstats_1$midPos-25000
sumstats_1$end=sumstats_1$midPos+25000
sumstats_1$chr <- gsub("chrLG", "", sumstats_1$chr)
sumstats_1$chr <- gsub("chrUn_", "", sumstats_1$chr)
sumstats_1$chr <- gsub("chr", "", sumstats_1$chr)
sumstats_1=GRanges(sumstats_1)

overlaps <- findOverlaps(background, sumstats_1)
df_background=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_background$category="Background"

overlaps <- findOverlaps(immune, sumstats_1)
df_immune=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_immune$category="Immunity"

overlaps <- findOverlaps(behav, sumstats_1)
df_behav=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_behav$category="Behaviour"

overlaps <- findOverlaps(locom, sumstats_1)
df_locom=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_locom$category="Locomotion"

overlaps <- findOverlaps(heat, sumstats_1)
df_heat=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_heat$category="Heat response"

overlaps <- findOverlaps(extended, sumstats_1)
df_extended=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_extended$category="GWAS Winchell"

overlaps <- findOverlaps(narrow, sumstats_1)
df_narrow=as.data.frame(sumstats_1[subjectHits(overlaps)])
df_narrow$category="GEA Winchell"

df_narrow=unique(df_narrow)
df_extended=unique(df_extended)
df_heat=unique(df_heat)
df_locom=unique(df_locom)
df_behav=unique(df_behav)
df_background=unique(df_background)
df_immune=unique(df_immune)

df=rbind(df_background,df_immune,df_behav,df_locom,df_heat,df_extended,df_narrow)

df$category <- factor(df$category, levels = c("Background"  ,      "Behaviour", "Heat response","Immunity" ,"Locomotion"  , "GWAS Winchell", "GEA Winchell"))

####Permutation test with chromosome circularization to test for significant differences in FST, pi, etc.
## median_FST=median(df_extended$Fst12)
## genome=reduce(sumstats_1,min.gapwidth=1000000000)
## for (i in 1:1000){
## A=GRanges(df_extended)
## perm=circularRandomizeRegions(A, genome=genome, max.retries=10,per.chromosome=TRUE)
## overlaps_perm <- findOverlaps(perm, sumstats_1)
## df_extended_perm=unique(as.data.frame(sumstats_1[subjectHits(overlaps_perm)]))
## median_FST=c(median_FST,median(df_extended_perm$Fst12))
## }
## pval_FST=length(subset(median_FST[-1],median_FST[-1]>median_FST[1]))/length(median_FST[-1])
```

We can also generate a Manhattan plot of the whole genome FST, and highlight the candidate regions we identified naively, as well as the few candidate genes/regions identified by Winchell.

```{r}
# Load required libraries
library(ggplot2)
library(ggrastr)
library(qvalue)
library(tidyr)
library(dplyr)
library(data.table)

gea_data=as.data.frame(fread("LFMM_scan_K3_total.txt"))
pval=colnames(gea_data)[c(3,4,6,8,10,12,14,16)]

gea_data$CHR <- gsub("chrLG", "", gea_data$CHR)
gea_data$CHR <- gsub("chrUn_", "", gea_data$CHR)
gea_data$CHR <- gsub("chr", "", gea_data$CHR)

gea_long <- gea_data %>%
  pivot_longer(cols = all_of(pval),
               names_to = "Test",
               values_to = "pval")

gea_long <- gea_long %>%
  group_by(Test) %>%
  mutate(qval = qvalue(pval)$qvalues) %>%
  ungroup()

gea_long <- gea_long %>%
  group_by(Test) %>%
  mutate(adj_P = p.adjust(pval, method="BH")) %>%
  ungroup()

ggplot(gea_long, aes(x = pval, color = factor(Test))) +
  geom_histogram(bin_width=0.1) + facet_grid(Test ~ .)

 


gea_data=subset(gea_long,gea_long$pval<0.01)

gea_data$CHR <- gsub("chrLG", "", gea_data$CHR)
gea_data$CHR <- gsub("chrUn_", "", gea_data$CHR)
gea_data$CHR <- gsub("chr", "", gea_data$CHR)

setDT(gea_data)

intervals_of_interest <- df_narrow
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)
# Use non-equi join
gea_data[, highlight := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight to "yes" for matching rows
gea_data[matches, highlight := "yes", on = .(CHR, POSITION)]

chr=c("4","2")
start=c(121495000,23410000)
end=c(123655000,25660000)
intervals_of_interest=as.data.frame(cbind(chr,start,end))

intervals_of_interest$chr=as.factor(intervals_of_interest$chr)
intervals_of_interest$start=as.integer(intervals_of_interest$start)
intervals_of_interest$end=as.integer(intervals_of_interest$end)


setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight2 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight2 to "yes" for matching rows
gea_data[matches, highlight2 := "yes", on = .(CHR, POSITION)]





intervals_of_interest <- df_extended
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight3 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight3 to "yes" for matching rows
gea_data[matches, highlight3 := "yes", on = .(CHR, POSITION)]



intervals_of_interest <- df_behav
colnames(intervals_of_interest)[1]="chr"

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight4 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight4 to "yes" for matching rows
gea_data[matches, highlight4 := "yes", on = .(CHR, POSITION)]




gea_data=subset(gea_data,(gea_data$CHR>0 & gea_data$CHR<7))


# Reshape gea_data to long format
desired_order <- c("P_val_combined", "env_PC1", "env_PC2", "env_PC3",
                   "urban_PC1", "canopy_cover", "nightlight", "per_impervious")

gea_data$Test <- factor(gea_data$Test, levels = desired_order)


pdf(file="GEA_Manhattan_pvalues.pdf",width=15,height=25)
ggplot(gea_data, aes(x = POSITION, y = -log10(adj_P), color = factor(CHR))) +
    geom_point_rast(size = 0.8) +
    scale_color_manual(values = rep(c("darkgrey", "black"), 16)) +
    geom_point_rast(data = subset(gea_data, 
                                  highlight == "yes"),
                    aes(x = POSITION, y = -log10(adj_P)), color = "darkgreen", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight2 == "yes"),
                    aes(x = POSITION, y = -log10(adj_P)), color = "red", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight3 == "yes"),
                    aes(x = POSITION, y = -log10(adj_P)), color = "darkorange", size = 1.5) +
#    geom_point_rast(data = subset(gea_data, highlight4 == "yes"),
#                    aes(x = POSITION, y = -log10(adj_P)), color = "lightblue", size = 1.5) +
    facet_grid(Test ~ CHR, scales = "free_y") +
    theme_minimal() +
    labs(x = "Genomic Position", y = "-log10(B-H adjusted p-value)", title = "Manhattan Plots by Test and Chromosome") +
    theme(legend.position = "none")
dev.off()



```

###Testing overlap between our study, scan for selection, and Winchell's study

We can then take candidate SNPs which pass the 1% Benjamini-Hochberg corrected P-value threshold. First, we look at whether these candidates fall in regions of higher PBS/Fst.

```{r}
#Find overlap
library(GenomicRanges)
library(IRanges)
library(data.table)
library(regioneR)
library(ggpubr)
library(biomaRt)
library(dplyr)

gea_long$outlier=ifelse(gea_long$adj_P<=0.01,paste("outlier",gea_long$Test,sep="_"),"neutral")
table(gea_long$outlier)

###make a Figure of PBS/Fst in windows overlapping an outlier


sumstats_1=as.data.frame(fread("Selection_tests/All_results_50kb_FST_PBS_ANGSD_countryside.txt",h=T))
sumstats_1$start=sumstats_1$midPos-25000
sumstats_1$end=sumstats_1$midPos+25000
sumstats_1$chr <- gsub("chrLG", "", sumstats_1$chr)
sumstats_1$chr <- gsub("chrUn_", "", sumstats_1$chr)
sumstats_1$chr <- gsub("chr", "", sumstats_1$chr)
sumstats_1=GRanges(sumstats_1)

names(table(gea_long$outlier))
#
  
outliers=subset(gea_long,gea_long$outlier != "neutral")
outliers$start=outliers$POSITION-1
outliers$end=outliers$POSITION

outliers=GRanges(outliers)
overlaps <- findOverlaps(outliers, sumstats_1)
fst_windows_outlier=as.data.frame(sumstats_1[subjectHits(overlaps)])
fst_windows_outlier$outlier_type <- outliers$outlier[queryHits(overlaps)]
fst_windows_outlier=unique(fst_windows_outlier)
# Indices of overlapping windows
overlap_indices <- subjectHits(overlaps)
# Logical vector of non-overlaps
non_overlap_indices <- setdiff(seq_along(sumstats_1), overlap_indices)
# Extract non-overlapping windows
fst_windows_non_outlier <- as.data.frame(sumstats_1[non_overlap_indices])
fst_windows_non_outlier$outlier_type="Background"

fst_windows=rbind(fst_windows_outlier,fst_windows_non_outlier)
boxplot(fst_windows_outlier$Fst12 ~ fst_windows_outlier$outlier_type)
boxplot(fst_windows_outlier$PBS2 ~  fst_windows_outlier$outlier_type)

desired_order <- c("Background", "outlier_P_val_combined" , "outlier_env_PC1" ,     "outlier_env_PC2"  ,      "outlier_env_PC3"  ,"outlier_urban_PC1"  ,"outlier_canopy_cover"   ,  "outlier_nightlight"    , "outlier_per_impervious" )

fst_windows$outlier_type <- factor(fst_windows$outlier_type, levels = desired_order)

pdf(file="Fst_PBS_GEA_outliers.pdf",width=14,height=12)

ggplot(fst_windows, aes(x = outlier_type , y = Fst12)) +
  geom_violin(fill="lightgrey") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme_minimal() +
  labs(title = "Differentiation between urban and non-urban anoles (LA)", x = "LFMM2 outlier test", y = "Fst")



ggplot(fst_windows, aes(x = outlier_type , y = PBS2)) +
  geom_violin(fill="lightgrey") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme_minimal() +
  labs(title = "Population-Branch-Statistic", x = "LFMM2 outlier test", y = "PBS test (urban v. others)")
dev.off()

### Upset plot for the GEA results and comparison with Winchell.
##We already have the Granges with gene coordinates custom_coords=GRanges(custom_coords)

outliers=subset(gea_long,gea_long$outlier != "neutral")
outliers$start=outliers$POSITION-50000  ###to make it similar to Fst window-based approach
outliers$end=outliers$POSITION+50000
outliers=GRanges(outliers)

overlaps <- findOverlaps(outliers,custom_coords)
genes_outlier=custom_coords[subjectHits(overlaps)]
genes_outlier$outlier_type <- outliers$outlier[queryHits(overlaps)]
genes_outlier=unique(as.data.frame(genes_outlier,row.names=seq(1,length(genes_outlier@ranges) )))

# Indices of overlapping windows
overlap_indices <- subjectHits(overlaps)
# Logical vector of non-overlaps
non_overlap_indices <- setdiff(seq_along(custom_coords), overlap_indices)
# Extract non-overlapping windows
genes_non_outlier <- as.data.frame(custom_coords[non_overlap_indices])
genes_non_outlier$outlier_type="Background"

genes=rbind(genes_outlier,genes_non_outlier)
length(unique(subset(genes,genes$category_strict_cristal==1 & genes$outlier_type !="Background")$gene))
###6 genes identified over 33.

length(unique(subset(genes,genes$category_strict_cristal==1 & genes$outlier_type =="outlier_urban_PC1")$gene))
#1 gene

genes_outlier$category_Fst_extreme=ifelse(genes_outlier$gene %in% c("ENSACAG00000016065","ENSACAG00000013324","ENSACAG00000013384"),1,0)
all_candidates=subset(genes_outlier,rowSums(genes_outlier[,c(7:12,14)])>0)
table(all_candidates$category_immune,all_candidates$outlier_type)
table(all_candidates$category_behaviour,all_candidates$outlier_type)
table(all_candidates$category_locomotion,all_candidates$outlier_type)
table(all_candidates$category_heat,all_candidates$outlier_type)
table(all_candidates$category_extended_cristal,all_candidates$outlier_type)
table(all_candidates$category_strict_cristal,all_candidates$outlier_type)



urban_candidates=subset(genes_outlier,rowSums(genes_outlier[,c(7:12,14)])>0 & genes_outlier$outlier_type %in% c("outlier_urban_PC1"  ,"outlier_canopy_cover"   ,  "outlier_nightlight"    , "outlier_per_impervious" ))
table(urban_candidates$category_immune,urban_candidates$outlier_type)
table(urban_candidates$category_behaviour,urban_candidates$outlier_type)
table(urban_candidates$category_locomotion,urban_candidates$outlier_type)
table(urban_candidates$category_heat,urban_candidates$outlier_type)
table(urban_candidates$category_extended_cristal,urban_candidates$outlier_type)
table(urban_candidates$category_strict_cristal,urban_candidates$outlier_type)

length(unique(urban_candidates$gene))  ###74 distinct genes.




urban_candidates_all=subset(genes_outlier, genes_outlier$outlier_type %in% c("outlier_urban_PC1"  ,"outlier_canopy_cover"   ,  "outlier_nightlight"    , "outlier_per_impervious" ))

length(unique(urban_candidates_all$gene))  ### distinct genes.

###We can output all candidate genes with their GO annotation if we have a list.


urban_candidates_unique <- urban_candidates %>%
  group_by(across(-outlier_type)) %>%  # Group by all columns except go_id
  summarise(outlier_type = paste(unique(outlier_type), collapse = ";"), .groups = "drop")

all_go_list_topdown=subset(anno,anno$external_gene_name%in%unique(urban_candidates$gene) | anno$ensembl_gene_id%in%unique(urban_candidates$gene))
all_go_list_topdown_unique <- all_go_list_topdown %>%
  group_by(across(-go_id)) %>%  # Group by all columns except go_id
  summarise(go_id = paste(unique(go_id), collapse = ";"), .groups = "drop")
colnames(urban_candidates_unique)[6]="ensembl_gene_id"
final=merge(urban_candidates_unique,all_go_list_topdown_unique,by="ensembl_gene_id")

write.table(final,file="Table_1_complement_GEA.txt",quote=F,sep="\t",row.names = F )

length(unique(final$ensembl_gene_id))  ### distinct genes.



all_candidates_unique <- all_candidates %>%
  group_by(across(-outlier_type)) %>%  # Group by all columns except go_id
  summarise(outlier_type = paste(unique(outlier_type), collapse = ";"), .groups = "drop")

all_go_list_topdown=subset(anno,anno$external_gene_name%in%unique(all_candidates$gene) | anno$ensembl_gene_id%in%unique(all_candidates$gene))
all_go_list_topdown_unique <- all_go_list_topdown %>%
  group_by(across(-go_id)) %>%  # Group by all columns except go_id
  summarise(go_id = paste(unique(go_id), collapse = ";"), .groups = "drop")
colnames(all_candidates_unique)[6]="ensembl_gene_id"
final=merge(all_candidates_unique,all_go_list_topdown_unique,by="ensembl_gene_id")

write.table(final,file="Sup_Table_GEA_topdown.txt",quote=F,sep="\t",row.names = F )

length(unique(final$ensembl_gene_id))  ### distinct genes.



gene_outlier_unique <- genes_outlier %>%
  group_by(across(-outlier_type)) %>%  # Group by all columns except go_id
  summarise(outlier_type = paste(unique(outlier_type), collapse = ";"), .groups = "drop")

all_go_list_topdown=subset(anno,anno$external_gene_name%in%unique(genes_outlier$gene) | anno$ensembl_gene_id%in%unique(genes_outlier$gene))
all_go_list_topdown_unique <- all_go_list_topdown %>%
  group_by(across(-go_id)) %>%  # Group by all columns except go_id
  summarise(go_id = paste(unique(go_id), collapse = ";"), .groups = "drop")
colnames(gene_outlier_unique)[6]="ensembl_gene_id"
final=merge(gene_outlier_unique,all_go_list_topdown_unique,by="ensembl_gene_id")

write.table(final,file="Sup_Table_all_GEA_outliers.txt",quote=F,sep="\t",row.names = F )

length(unique(final$ensembl_gene_id))  ### distinct genes.


```

We can test whether the overlap we find between candidate genes and outlier SNPs is greater than expected by chance, taking gene length and relative distances into account:

```{r}

 ####Permutation test with chromosome circularization to test for significant differences in outlier overlap
library(BiocParallel)



n_outliers_GEA = genes_outlier[,1:(ncol(genes_outlier)-1)]
n_outliers_GEA$replicate=0 ##observed
genome=GenomicRanges::reduce(sumstats_1,min.gapwidth=1000000000)
A=GRanges(outliers)

 for (i in 1:10000){
 perm=circularRandomizeRegions(A, genome=genome, max.retries=10,per.chromosome=TRUE)
#perm=circularRandomizeRegions(A, genome=genome, max.retries=10,per.chromosome=FALSE)
print(i)
overlaps_perm <- findOverlaps(perm, custom_coords)
perm_outlier=custom_coords[subjectHits(overlaps_perm)]
perm_outlier$outlier_type <- perm$outlier[queryHits(overlaps_perm)]
perm_outlier=unique(as.data.frame(perm_outlier,row.names=seq(1,length(perm_outlier@ranges) )))
perm_outlier$replicate=i
n_outliers_GEA=rbind(n_outliers_GEA,perm_outlier)
 }


library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

long_data <- n_outliers_GEA %>%
  pivot_longer(
    cols = starts_with("category_"),
    names_to = "category",
    values_to = "in_category"
  ) %>%
  filter(in_category == 1)  # keep only genes in that category


category_counts <- long_data %>%
  group_by(outlier_type, replicate, category) %>%
  summarize(
    n_genes = n_distinct(gene),
    .groups = "drop"
  )

pdf(file="Permutation_tests_GEA_threshold1pc.pdf",width=18,height = 12)
ggplot(category_counts %>% filter(replicate != 0), aes(x = n_genes)) +
  geom_histogram(binwidth = 1, fill = "grey80", color = "black") +
  geom_vline(data = category_counts %>% filter(replicate == 0),
             aes(xintercept = n_genes),
             color = "red", linewidth = 1) +
  facet_grid(category ~ outlier_type, scales = "free_y") +
  theme_minimal() +
  labs(title = "Permutation distribution of category overlaps",
       subtitle = "Red line = observed (replicate 0)",
       x = "# overlapping category genes", y = "Permutation frequency")

dev.off()
pval_table <- category_counts %>%
  group_by(outlier_type, category) %>%
  summarize(
    observed = n_genes[replicate == 0],
    permuted = list(n_genes[replicate != 0]),
    p_value = mean(n_genes[replicate != 0] >= n_genes[replicate == 0]),
    .groups = "drop"
  )
write.table(as.data.frame(pval_table[,c(1,2,3,5)]),file="permutation_GEA.txt",quote=F,row.names=F,sep="\t")
##only significant category at adj_p=0.01 is locomotion 




 
 ###Significant overlap with all urban outlier tests at once?
urban_candidates=subset(genes_outlier, genes_outlier$outlier_type %in% c("outlier_urban_PC1"  ,"outlier_canopy_cover"   ,  "outlier_nightlight"    , "outlier_per_impervious" ))



n_outliers_GEA = urban_candidates[,1:(ncol(urban_candidates)-1)]
n_outliers_GEA$outlier_type="all urban tests"
n_outliers_GEA$replicate=0 ##observed
n_outliers_GEA=unique(n_outliers_GEA)
genome=GenomicRanges::reduce(sumstats_1,min.gapwidth=1000000000)
A=GRanges(outliers)

 for (i in 1:100){
 perm=circularRandomizeRegions(A, genome=genome, max.retries=10,per.chromosome=TRUE)
#perm=circularRandomizeRegions(A, genome=genome, max.retries=10,per.chromosome=FALSE)

overlaps_perm <- findOverlaps(perm, custom_coords)
perm_outlier=custom_coords[subjectHits(overlaps_perm)]
perm_outlier$outlier_type <- perm$outlier[queryHits(overlaps_perm)]
perm_outlier=subset(perm_outlier, perm_outlier$outlier_type %in% c("outlier_urban_PC1"  ,"outlier_canopy_cover"   ,  "outlier_nightlight"    , "outlier_per_impervious" ))
perm_outlier=unique(as.data.frame(perm_outlier,row.names=seq(1,length(perm_outlier@ranges) )))
perm_outlier$outlier_type="all urban tests"
perm_outlier=unique(perm_outlier)
perm_outlier$replicate=i
n_outliers_GEA=rbind(n_outliers_GEA,perm_outlier)
 }


library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

long_data <- n_outliers_GEA %>%
  pivot_longer(
    cols = starts_with("category_"),
    names_to = "category",
    values_to = "in_category"
  ) %>%
  filter(in_category == 1)  # keep only genes in that category


category_counts <- long_data %>%
  group_by(outlier_type, replicate, category) %>%
  summarize(
    n_genes = n_distinct(gene),
    .groups = "drop"
  )

pdf(file="Permutation_tests_GEA_threshold1pc_urban_combined.pdf",width=18,height = 12)
ggplot(category_counts %>% filter(replicate != 0), aes(x = n_genes)) +
  geom_histogram(binwidth = 1, fill = "grey80", color = "black") +
  geom_vline(data = category_counts %>% filter(replicate == 0),
             aes(xintercept = n_genes),
             color = "red", linewidth = 1) +
  facet_grid(category ~ outlier_type, scales = "free_y") +
  theme_minimal() +
  labs(title = "Permutation distribution of category overlaps",
       subtitle = "Red line = observed (replicate 0)",
       x = "# overlapping category genes", y = "Permutation frequency")

dev.off()
pval_table <- category_counts %>%
  group_by(outlier_type, category) %>%
  summarize(
    observed = n_genes[replicate == 0],
    permuted = list(n_genes[replicate != 0]),
    p_value = mean(n_genes[replicate != 0] >= n_genes[replicate == 0]),
    .groups = "drop"
  )
write.table(as.data.frame(pval_table[,c(1,2,3,5)]),file="permutation_GEA_urban_combined.txt",quote=F,row.names=F,sep="\t")
##only significant category at adj_p=0.01 is locomotion 


##perm_fn <- function(i) {
##    tryCatch({
##perm <- circularRandomizeRegions(A, genome = genome, max.retries = 10, per.chromosome = TRUE)
##  overlaps_perm <- findOverlaps(perm, custom_coords)
##  perm_outlier <- custom_coords[subjectHits(overlaps_perm)]
##  perm_outlier$outlier_type <- perm$outlier[queryHits(overlaps_perm)]
##  df <- as.data.frame(perm_outlier,row.names=NULL)
##  df$replicate <- i
##  df}, error = function(e) {
##    warning(sprintf("Permutation %d failed: %s", i, e$message))
##    NULL
##  })
##}
### Set up parallel backend
##param <- MulticoreParam(workers = 4)  # Adjust number of cores as needed
##
### Run permutations in parallel
##n_perm <- 1000
##perm_list <- bplapply(seq_len(n_perm), perm_fn, BPPARAM = param)
##
### Combine all results
##n_outliers_GEA <- bind_rows(n_outliers_GEA, do.call(rbind, perm_list))
##





###GO term enrichment analysis, extract candidate genes overlapping with GEA, GWAS, behaviour, etc.

###Last panel could be a map of structure for combined outliers, or urban PC, let's see. Sup Material could show for each category of outlier. requires filtering + PCA with plink but not too difficult.



```

At last, we can inspect the distribution of p-values in candidate genes v. permuted datasets: for example, do we find lower p-values for genes ranked in the lowest 5% of their category compared to permuted datasets? What about the median?

```{r}

library(GenomicRanges)
library(GenomicInteractions)
library(dplyr)
library(tidyr)
library(regioneR)  # for circularRandomizeRegions
library(ggplot2)
library(data.table)
library(GenomicRanges)
library(IRanges)
library(regioneR)
library(dplyr)
library(BiocParallel)

# Detect cores and set up a parallel backend
param <- MulticoreParam(workers = 4)  # Adjust number of cores to your system


# --- Input ---
genes <- custom_coords  # gene annotations with category columns
genome <- reduce(genes, min.gapwidth = 1e9)
outliers <- gea_data

# --- Step 1: Define tests ---
tests <- grep("^canopy_cover|^env_PC|^nightlight|^per_impervious|^urban_PC1|^P_val_combined", 
               names(outliers), value = TRUE)

# --- Step 2: Annotate each SNP with its 50kb window ---
outliers$window_start <- floor(outliers$POSITION / 50000) * 50000
outliers$window_chr <- outliers$CHR
outliers$window_end <- outliers$window_start + 49999

outliers$CHR_0 <- paste0(outliers$CHR, "_", outliers$window_start)

# --- Step 3: Minimum p-values per window for observed data ---
get_min_p <- function(test) {
  df <- outliers[, c("CHR_0", test)]
  colnames(df) <- c("window", "pval")
  df %>%
    group_by(window) %>%
    summarize(min_pval = min(pval, na.rm = TRUE), .groups = "drop") %>%
    mutate(test = test)
}
observed_df <- bind_rows(lapply(tests, get_min_p))
observed_df$replicate <- 0

# --- Step 4: Map windows to coordinates ---
window_coords <- outliers %>% 
  dplyr::select(CHR_0, window_chr, window_start, window_end) %>% 
  distinct()

window_gr <- GRanges(seqnames = window_coords$window_chr,
                     ranges = IRanges(start = window_coords$window_start, 
                                      end = window_coords$window_end),
                     CHR_0 = window_coords$CHR_0)

# --- Step 5: Rotate genes instead of SNPs ---
genes_gr <- genes  # GRanges with gene coords + category columns

n_perm <- 100
all_perms <- vector("list", n_perm)
#perm_list <- bplapply(seq_len(n_perm), function(i) {

for (i in seq_len(n_perm)) {
  message("Permutation ", i, " of ", n_perm)

  #genome_subset <- genome[seqnames(genome) %in% seqlevelsInUse(genes_gr)]
  # Permute gene positions
  permuted_genes <- circularRandomizeRegions(genes_gr, genome = genome, per.chromosome = TRUE)
  #mcols(permuted_genes) <- mcols(genes_gr)  # retain categories

  # Find overlaps with windows
  ov <- findOverlaps(permuted_genes, window_gr)
  overlapping_windows <- window_gr[subjectHits(ov)]
  gene_ids <- queryHits(ov)

# Build a data frame of overlapping windows and associated gene IDs
overlap_df <- data.frame(
  gene_id = gene_ids,
  CHR_0 = overlapping_windows$CHR_0,
  stringsAsFactors = FALSE
)

# Create all combinations of gene-window and test
expanded_df <- merge(overlap_df, data.frame(test = tests), by = NULL)

# Join with the p-values
min_pvals <- observed_df %>% 
  rename(CHR_0 = window)

joined <- left_join(expanded_df, min_pvals, by = c("CHR_0", "test"))
joined$replicate <- i
#return(joined)},BPPARAM=param) ##and comment out below
all_perms[[i]] <- joined

}

#all_perms <- bind_rows(perm_list) ##comment to remove for parallel

perm_df <- bind_rows(all_perms)
perm_df <- perm_df %>% 
  rename(min_pval = min_pval)

# --- Step 6: Get observed values (attach window -> gene once) ---
ov_obs <- findOverlaps(genes_gr, window_gr)
gene_ids_obs <- queryHits(ov_obs)
window_chr0_obs <- window_gr$CHR_0[subjectHits(ov_obs)]

obs_df <- expand.grid(gene_id = gene_ids_obs, test = tests, stringsAsFactors = FALSE)
obs_df$CHR_0 <- rep(window_chr0_obs, times = length(tests))
obs_df <- left_join(obs_df, observed_df, by = c("CHR_0" = "window", "test"))
obs_df$replicate <- 0

# --- Step 7: Combine observed and permuted ---
combined <- bind_rows(obs_df, perm_df)
combined$gene_id <- as.integer(combined$gene_id)

# --- Step 8: Merge gene categories ---
gene_cat <- as.data.frame(mcols(genes_gr))
gene_cat$gene_id <- seq_len(nrow(gene_cat))
long_combined <- left_join(combined, gene_cat, by = "gene_id")

category_cols <- grep("^category_", names(long_combined), value = TRUE)

# --- Step 9: Summarize per category ---
summary_list <- lapply(category_cols, function(cat) {
  df <- long_combined %>% filter(.data[[cat]] == 1)
  df %>%
    group_by(test, replicate) %>%
    summarize(
      across(
        .cols = min_pval,
        .fns = list(
          q05 = ~ quantile(., probs = 0.05, na.rm = TRUE),
          q10 = ~ quantile(., probs = 0.10, na.rm = TRUE),
          q25 = ~ quantile(., probs = 0.25, na.rm = TRUE),
          q50 = ~ quantile(., probs = 0.50, na.rm = TRUE),
          q75 = ~ quantile(., probs = 0.75, na.rm = TRUE),
          q90 = ~ quantile(., probs = 0.90, na.rm = TRUE),
          q95 = ~ quantile(., probs = 0.95, na.rm = TRUE)
        )
      ),
      .groups = "drop"
    ) %>%
    mutate(category = cat)
  })
summary_df <- bind_rows(summary_list)


# Define quantiles of interest
quantiles <- c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)



summary_df <- bind_rows(summary_list)

long_df <- summary_df %>%
  pivot_longer(cols = starts_with("min_pval_"), names_to = "quantile", values_to = "value")

# Split observed and permuted
observed_vals <- long_df %>% filter(replicate == 0)
permuted_vals <- long_df %>% filter(replicate != 0)

# Calculate empirical p-values
pval_summary <- observed_vals %>%
  rename(observed = value) %>%
  left_join(
    permuted_vals %>%
      group_by(test, category, quantile) %>%
      summarize(pval = mean(value < observed_vals$value[observed_vals$test == test &
                                                         observed_vals$category == category &
                                                         observed_vals$quantile == quantile], na.rm = TRUE),
                .groups = "drop"),
    by = c("test", "category", "quantile")
  )

print(pval_summary)
write.table(pval_summary,file="permutation_test_distributions.txt",quote=F,sep="\t",row.names=F)

# Plot observed vs permuted
library(ggplot2)
ggplot(category_summary, aes(x = replicate, y = -log10(median_p), color = replicate == 0)) +
  geom_boxplot(aes(group = replicate)) +
  facet_wrap(~Category + Test, scales = "free_y") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red")) +
  labs(title = "Median -log10(p) per gene category/test across permutations")

```

### GO term enrichment analysis

We also run two more GO term enrichment analysis: one based on the multinomial test on env_PC + urban_PC, and the other based specifically on urban association (found significant in urban_PC1, per_impervious, nightlight, and canopy cover).

```{r}


library(biomaRt)
library(dplyr)
library(GOfuncR)

ensembl <- useMart("ensembl",host="https://may2024.archive.ensembl.org", dataset = "acarolinensis_gene_ensembl")  ##This was the version used for the paper, latest one is weird with missing GO terms
getBM(mart=ensembl,
      attributes=c('ensembl_gene_id','external_gene_name', 'go_id', 'name_1006'),
      filters="external_gene_name",
      values="Tl")
anno <- getBM(mart=ensembl,attributes=c("ensembl_gene_id", "external_gene_name", "go_id", "chromosome_name", "start_position", "end_position", "strand", "description"))

go_annot=anno[,c(1,3)]
colnames(go_annot)=c("gene","go_id")
custom_coords=anno[,c(1,4,5,6)]
custom_coords$chromosome_name=gsub('.1','',custom_coords$chromosome_name)
colnames(custom_coords)=c("gene","chr","start","end")

###use gea_long/outliers object  instead

all_regions=gea_long
all_regions$start=all_regions$POSITION-50000
all_regions$end=all_regions$POSITION+50000
All=GenomicRanges::reduce(GRanges(all_regions),min.gapwidth=1000000000)

all_regions$is_candidate=ifelse(all_regions$outlier=="outlier_P_val_combined",1,0)
#all_regions$is_candidate=ifelse(all_regions$outlier %in% c("outlier_canopy_cover", "outlier_nightlight","outlier_per_impervious","outlier_urban_PC1",1,0)
                                
candidateregions=GenomicRanges::reduce(GRanges(subset(all_regions,all_regions$is_candidate==1)),min.gapwidth=1) ###We basically merge outliers if found at < 50kb.
backgroundregions=GenomicRanges::setdiff(All, candidateregions)
  
candidateregions=as.data.frame(candidateregions)
backgroundregions=as.data.frame(backgroundregions)
backgroundregions$start=ifelse(backgroundregions$start<1,1,backgroundregions$start)
candidateregions$start=ifelse(candidateregions$start<1,1,candidateregions$start)

candidateregions$regions=paste(paste(candidateregions$seqnames,candidateregions$start,sep=":"),candidateregions$end,sep="-")
candidateregions$candidate=1
backgroundregions$regions=paste(paste(backgroundregions$seqnames,backgroundregions$start,sep=":"),backgroundregions$end,sep="-")
backgroundregions$candidate=0
all_regions=rbind(backgroundregions,candidateregions)
all_regions=all_regions[order(all_regions[,1],all_regions[,2]),]
all_regions=all_regions[,c(6,7)]
###use genes object  instead

res_region = go_enrich(all_regions, n_randsets=100, annotations=go_annot, gene_coords=custom_coords, regions=TRUE,circ_chrom=T)
stats_region = res_region[[1]]
head(stats_region)
top_gos_hyper = subset(stats_region,stats_region$raw_p_overrep<0.01 & stats_region$ontology=="biological_process")[, 'node_id']
plot_anno_scores(res_region, top_gos_hyper[1:10], annotations=go_annot)
Table3=subset(stats_region,stats_region$raw_p_overrep<=0.01 & stats_region$ontology=="biological_process")
write.table(Table3,file="Table_3.txt",quote=F,sep="\t",row.names = F)





###Urban specific test

all_regions=gea_long
all_regions$start=all_regions$POSITION-50000
all_regions$end=all_regions$POSITION+50000
all_regions$is_candidate=ifelse(all_regions$outlier %in% c("outlier_canopy_cover", "outlier_nightlight","outlier_per_impervious","outlier_urban_PC1"),1,0)
                                
candidateregions=GenomicRanges::reduce(GRanges(subset(all_regions,all_regions$is_candidate==1)),min.gapwidth=1) ###We basically merge outliers if found at < 50kb.
backgroundregions=GenomicRanges::setdiff(All, candidateregions)
  
candidateregions=as.data.frame(candidateregions)
backgroundregions=as.data.frame(backgroundregions)
backgroundregions$start=ifelse(backgroundregions$start<1,1,backgroundregions$start)
candidateregions$start=ifelse(candidateregions$start<1,1,candidateregions$start)

candidateregions$regions=paste(paste(candidateregions$seqnames,candidateregions$start,sep=":"),candidateregions$end,sep="-")
candidateregions$candidate=1
backgroundregions$regions=paste(paste(backgroundregions$seqnames,backgroundregions$start,sep=":"),backgroundregions$end,sep="-")
backgroundregions$candidate=0
all_regions=rbind(backgroundregions,candidateregions)
all_regions=all_regions[order(all_regions[,1],all_regions[,2]),]
all_regions=all_regions[,c(6,7)]
###use genes object  instead

res_region = go_enrich(all_regions, n_randsets=100, annotations=go_annot, gene_coords=custom_coords, regions=TRUE,circ_chrom=T)
stats_region = res_region[[1]]
head(stats_region)
top_gos_hyper = subset(stats_region,stats_region$raw_p_overrep<0.01 & stats_region$ontology=="biological_process")[, 'node_id']
plot_anno_scores(res_region, top_gos_hyper[1:10], annotations=go_annot)
Table4=subset(stats_region,stats_region$raw_p_overrep<=0.01 & stats_region$ontology=="biological_process")
write.table(Table4,file="Table_4.txt",quote=F,sep="\t",row.names = F)



```

###Figure 5

```{r}

library(ComplexUpset)
library(dplyr)
library(tidyr)
library(scales)

urban=subset(gea_data,gea_data$Test == "urban_PC1" | gea_data$Test =="P_val_combined")
highlight_regions <- data.frame(
  CHR = rep(c(2, 4, 2, 1),2),
  Test = rep(c("P_val_combined", "urban_PC1"), each = 4),  # or adjust based on which panel(s) should show what
  xmin = rep(c(24310000, 121495000, 70500000, 114000000),2),
  xmax = rep(c(24760000, 123655000, 72500000, 125000000),2),
  ymin = -Inf,
  ymax = Inf,
  region_label = rep(c("Fst peak (chr2)", "Fst peak (chr4)", "Hox cluster", "Winchell GEA"),2)
)
panelA=ggplot(urban, aes(x = POSITION, y = -log10(adj_P), color = factor(CHR))) +
  geom_point_rast(size = 0.8) +
  scale_color_manual(values = rep(c("darkgrey", "black"), 16))  +
  geom_point_rast(data = subset(urban, highlight2 == "yes"),
                  aes(x = POSITION, y = -log10(adj_P)), color = "red", size = 1.5) +
  geom_point_rast(data = subset(urban, highlight3 == "yes"),
                  aes(x = POSITION, y = -log10(adj_P)), color = "darkorange", size = 1.5) +
  geom_point_rast(data = subset(urban, highlight == "yes"),
                  aes(x = POSITION, y = -log10(adj_P)), color = "darkgreen", size = 1.5) +
  # Add the rectangles for the regions
geom_rect(data = highlight_regions, 
          aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), 
          fill = "orange", alpha = 0.2, color = "black",
          inherit.aes = FALSE) +  facet_grid(Test ~ CHR, scales = "free_y") +
  theme_bw() +
  labs(x = "Genomic Position", y = "-log10(B-H adjusted p-value)", title = "") +
  theme(legend.position = "none")


##We want to show chr2, chr4, Winchell GEA chr1 and HOX on chr2. All Manhattan plots as Sup Mat, and take P_value combined and urban_PC1 for the main Figure 5 (easier and does a better job)

zooms=subset(gea_data,(gea_data$CHR==2 & gea_data$POSITION> (24410000 - 1e6) & gea_data$POSITION < (24660000 + 1e6)) | (gea_data$CHR==4 & gea_data$POSITION> (122495000 - 1e6) & gea_data$POSITION < (122655000 + 1e6)) | (gea_data$CHR==2 & gea_data$POSITION> 70.5e6 & gea_data$POSITION < 72.5e6) | (gea_data$CHR==1 & gea_data$POSITION> 114e6 & gea_data$POSITION < 125e6) )
zooms$region=ifelse(zooms$CHR==4,"Top Fst outlier region, chr4","NA")
zooms$region=ifelse(zooms$CHR==2 & zooms$POSITION<50e6,"Top Fst outlier region, chr2",zooms$region)
zooms$region=ifelse(zooms$CHR==2 & zooms$POSITION>50e6,"Hox genes region, chr2",zooms$region)
zooms$region=ifelse(zooms$CHR==1,"Winchell's GEA candidate region, chr1",zooms$region)

zooms <- zooms %>%
  mutate(highlight = ifelse(-log10(adj_P) > 2, "highlight", "normal"))


# Plot with conditional coloring
ggplot(zooms, aes(x = POSITION, y = -log10(adj_P), color = highlight)) +
  geom_point_rast(size = 0.8) +
  scale_color_manual(values = c("highlight" = "red", "normal" = "darkgrey")) +
  facet_grid(Test ~ region, scales = "free") +
  theme_bw() +
  labs(x = "Genomic Position", y = "-log10(B-H adjusted p-value)", title = "") +
  theme(legend.position = "none")

###Panel B

zooms=subset(urban,(urban$CHR==2 & urban$POSITION> (24410000 - 1e6) & urban$POSITION < (24660000 + 1e6)) | (urban$CHR==4 & urban$POSITION> (122495000 - 1e6) & urban$POSITION < (122655000 + 1e6)) | (urban$CHR==2 & urban$POSITION> 70.5e6 & urban$POSITION < 72.5e6) | (urban$CHR==1 & urban$POSITION> 114e6 & urban$POSITION < 125e6) )

zooms$region=ifelse(zooms$CHR==4,"4 Top Fst outlier region, chr4","NA")
zooms$region=ifelse(zooms$CHR==2 & zooms$POSITION<50e6,"3 Top Fst outlier region, chr2",zooms$region)
zooms$region=ifelse(zooms$CHR==2 & zooms$POSITION>50e6,"2 Hox genes region, chr2",zooms$region)
zooms$region=ifelse(zooms$CHR==1,"1 Winchell's GEA candidate region, chr1",zooms$region)

library(GenomicRanges)
library(dplyr)




snp_gr <- GRanges(seqnames = zooms$CHR,
                  ranges = IRanges(start = zooms$POSITION, width = 1))

# --- Step 2: Filter and convert genes to GRanges
final_2 <- subset(final, outlier_type %in% c("outlier_P_val_combined", "outlier_urban_PC1"))

gene_gr <- GRanges(seqnames = final_2$seqnames,
                   ranges = IRanges(start = final_2$start, end = final_2$end),
                   gene = final_2$ensembl_gene_id,
                   ext_name = final_2$external_gene_name)
gene_gr_expanded <- resize(gene_gr, width = width(gene_gr) + 50000, fix = "center")
gene_gr=gene_gr_expanded
# --- Step 3: Find overlaps and annotate SNPs
hits <- findOverlaps(snp_gr, gene_gr)
zooms$outlier_gene <- "normal"
zooms$outlier_gene[queryHits(hits)] <- "outlier"

zooms$gene <- NA
zooms$gene[queryHits(hits)] <- mcols(gene_gr)$gene[subjectHits(hits)]
zooms$ext_name <- NA
zooms$ext_name[queryHits(hits)] <- mcols(gene_gr)$ext_name[subjectHits(hits)]
#zooms$ext_name<-ifelse(zooms$ext_name=="",zooms$gene,zooms$ext_name) #too many unnamed genes, we just show the ones with an external name.
# --- Step 4: Create label dataframe per Test per gene
label_df <- zooms %>%
  filter(outlier_gene == "outlier") %>%
  filter(!is.na(gene)) %>%
  group_by(Test, gene) %>%
  slice_min(order_by = adj_P, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(-log10(adj_P) > 2)  # Only label significant ones

zooms$highlight <- ifelse(-log10(zooms$adj_P) > 2, "highlight", "normal")

# --- Step 5: Plot

panelB=ggplot(zooms, aes(x = POSITION, y = -log10(adj_P))) +
geom_point_rast(aes(color = highlight), size = 0.8) +
scale_color_manual(values = c("normal" = "darkgrey", "highlight" = "orange")) +
  geom_text_repel(data = label_df,
                  aes(label = ext_name),
                  size = 3.5,   fontface = "bold",
max.overlaps = 50, segment.size = 0.5) +
  facet_grid(Test ~ region, scales = "free") + scale_x_continuous(labels = label_scientific()) +
  theme_bw() +
  labs(x = "Genomic Position", y = "-log10(B-H adjusted p-value)") +
  theme(legend.position = "none")


library(magick)
library(grid)
library(gridExtra)
pdf(file="GEA_Figure5.pdf",width=22,height=10)
ggarrange(panelA, panelB,labels = c("A", "B"),font.label = list(size = 28, color = "black", face = "bold"),ncol = 2)
dev.off()


##to generate an upset plot with genes, we have a file with 2 columns named sharing_outlier_tests.txt

gene_outlier_unique <- genes_outlier 

all_go_list_topdown=subset(anno,anno$external_gene_name%in%unique(genes_outlier$gene) | anno$ensembl_gene_id%in%unique(genes_outlier$gene))
all_go_list_topdown_unique <- all_go_list_topdown %>%
  group_by(across(-go_id)) %>%  # Group by all columns except go_id
  summarise(go_id = paste(unique(go_id), collapse = ";"), .groups = "drop")
colnames(gene_outlier_unique)[6]="ensembl_gene_id"
final=merge(gene_outlier_unique,all_go_list_topdown_unique,by="ensembl_gene_id")

last=final[,c(which(colnames(final)=="ensembl_gene_id"),which(colnames(final)=="outlier_type"))]
last=subset(last,last$outlier_type !="outlier_P_val_combined")
binary_matrix <- last %>%
  distinct() %>%  # Remove any accidental duplicates
  mutate(value = 1) %>%
  pivot_wider(names_from = outlier_type, values_from = value, values_fill = list(value = 0))

# Step 2: Make UpSet plot
upset(
  binary_matrix,
  intersect = colnames(binary_matrix)[-1],  # All columns except gene ID
  base_annotations = list(
    'Intersection size' = intersection_size()
  ),
  name = "Outlier sets"
)

```

### Further check FST in How genes region

```{r}


###Check Fst in the large chr2 block overlapping Hox genes
sumstats1=as.data.frame(fread("Selection_tests/All_results_5kb_FST_PBS_ANGSD_countryside.txt",h=T))

hox_coords=anno[grep("HOX", anno$external_gene_name),]
hox_coords=unique(subset(hox_coords,hox_coords$chromosome_name==2))
Hox=subset(sumstats1,sumstats1$chr=="chr2")
Hox=Hox[order(Hox$midPos),]
par(mfrow=c(2,2))
plot(Hox$midPos,Hox$Fst12,xlim=c(55e6,83e6),xlab="Position",ylab="Fst",type="l")
plot(Hox$midPos,Hox$PBS2,xlim=c(55e6,83e6),xlab="Position",ylab="PBS",type="l")

Hox$highlight <- FALSE

for (i in 1:nrow(hox_coords)) {
  Hox$highlight <- Hox$highlight | 
    (Hox$midPos >= hox_coords$start_position[i] & 
     Hox$midPos <= hox_coords$end_position[i])
}

A=ggplot(Hox, aes(x = midPos, y = Fst12)) +
  geom_point(color = "grey60", size = 0.7) +
  # Then, overlay points in red where highlight == TRUE
  geom_point(data = subset(Hox, highlight == TRUE),
             aes(x = midPos, y = Fst12),
             color = "red", size = 2) +
  theme_minimal() +
  labs(x = "Genomic position", y = "Fst") +
  theme(panel.grid.minor = element_blank()) + xlim(55e6,83e6) 



# Example plot for PBS2
B=ggplot(Hox, aes(x = midPos, y = PBS2)) +
  geom_point(color = "grey60", size = 0.7) +
  # Then, overlay points in red where highlight == TRUE
  geom_point(data = subset(Hox, highlight == TRUE),
             aes(x = midPos, y = PBS2),
             color = "red", size = 2) +
  theme_minimal() +
  labs(x = "Genomic position", y = "PBS") +
  theme(panel.grid.minor = element_blank()) + xlim(55e6,83e6) 




sumstats1=as.data.frame(fread("Selection_tests/All_results_50kb_FST_PBS_ANGSD_countryside.txt",h=T))
Hox=subset(sumstats1,sumstats1$chr=="chr2")
Hox=Hox[order(Hox$midPos),]
plot(Hox$midPos,Hox$Fst12,xlim=c(55e6,83e6),xlab="Position",ylab="Fst",type="l")
plot(Hox$midPos,Hox$PBS2,xlim=c(55e6,83e6),xlab="Position",ylab="PBS",type="l")


Hox$highlight <- FALSE

for (i in 1:nrow(hox_coords)) {
  Hox$highlight <- Hox$highlight | 
    (Hox$midPos >= hox_coords$start_position[i] & 
     Hox$midPos <= hox_coords$end_position[i])
}

# Example plot for Fst12
C=ggplot(Hox, aes(x = midPos, y = Fst12)) +
  geom_point(color = "grey60", size = 0.7) +
  # Then, overlay points in red where highlight == TRUE
  geom_point(data = subset(Hox, highlight == TRUE),
             aes(x = midPos, y = Fst12),
             color = "red", size = 2) +
  theme_minimal() +
  labs(x = "Genomic position", y = "Fst") +
  theme(panel.grid.minor = element_blank()) + xlim(55e6,83e6) 



# Example plot for PBS2
D=ggplot(Hox, aes(x = midPos, y = PBS2)) +
  geom_point(color = "grey60", size = 0.7) +
  # Then, overlay points in red where highlight == TRUE
  geom_point(data = subset(Hox, highlight == TRUE),
             aes(x = midPos, y = PBS2),
             color = "red", size = 2) +
  theme_minimal() +
  labs(x = "Genomic position", y = "PBS") +
  theme(panel.grid.minor = element_blank()) + xlim(55e6,83e6) 

ggarrange(A,B,C,D)




```

### PCA and map of outliers (urban outliers v. P_value combined outliers)

First we need to extract all candidate SNPs.

```{r}

urban=unique(subset(gea_long,gea_long$adj_P<=0.01 & gea_long$Test  %in% c("urban_PC1"  ,"canopy_cover"   ,  "nightlight"    , "per_impervious" ))[,c(1,2,2)])
urban$rangeID=seq(1,nrow(urban))
multinomial=unique(subset(gea_long,gea_long$adj_P<=0.01 & gea_long$Test  %in% c("P_val_combined" ))[,c(1,2,2)])
multinomial$rangeID=seq(1,nrow(multinomial))

##6740 unique variants for urban
##33806 unique variants for multinomial
write.table(urban,file="urban_SNPs_positions_outliers.txt",sep="\t",col.names = F,row.names = F,quote=F)
write.table(multinomial,file="multinomial_SNPs_positions_outliers.txt",sep="\t",col.names = F,row.names = F,quote=F)

```

In bash, to also obtain non-synonymous variants, if any.

```{bash}

sed -i -e "s:GL:chrUn_GL:" urban_SNPs_positions_outliers.txt
sed -i -e "s:^c\t:chrLGc\t:" urban_SNPs_positions_outliers.txt
sed -i -e "s:^a:chrLGa:" urban_SNPs_positions_outliers.txt
sed -i -e "s:^b:chrLGb:" urban_SNPs_positions_outliers.txt

grep CDS Anolis_carolinensis.AnoCar2.0v2.114.gtf | sed -E 's/^([^[:space:]]+)\.1(\t)/\1\2/'  | sed "s:^GL:chrUn_GL:" | sed "s:^c\t:chrLGc\t:" | sed  "s:^a:chrLGa:" | sed "s:^b:chrLGb:"  > CDS_positions.gtf

sed -e "s:^1\t:chr1\t:" urban_SNPs_positions_outliers.txt |sed -e "s:^2\t:chr2\t:" | sed -e "s:^3\t:chr3\t:" | sed -e "s:^4\t:chr4\t:" |
sed -e "s:^5\t:chr5\t:" | sed -e "s:^6\t:chr6\t:" > urban_SNPs_forvcftools.txt

module load vcftools
vcftools --gzvcf All_imputed.vcf.gz --positions urban_SNPs_forvcftools.txt --recode --recode-INFO-all --out urban_SNPs

module load snpeff/
cp /shared/ifbstor1/software/miniconda/envs/snpeff-5.2/share/snpeff-5.2-0/snpEff.config .
cp  urban_SNPs.recode.vcf save.vcf
sed -i -e "s:chrUn_::"  urban_SNPs.recode.vcf
sed -i -e "s:chrLG::"  urban_SNPs.recode.vcf
sed -i -e "s:chr::"  urban_SNPs.recode.vcf
sed -i -E 's/^(AA[^[:space:]]*|GL[^[:space:]]*)/\1.1/'  urban_SNPs.recode.vcf

mkdir data
snpEff -c snpEff.config -v AnoCar2.0v2.105 urban_SNPs.recode.vcf -V > urban_ann.vcf

grep MODERATE urban_ann.vcf | cut -f8 | cut -f7 -d ";" | cut -f4 -d "|" | sort | uniq 
###13 unique genes
grep HIGH urban_ann.vcf | cut -f8 | cut -f7 -d ";" | cut -f4 -d "|" | sort | uniq 
##1 unique gene

grep MODERATE urban_ann.vcf | cut -f8 | cut -f7 -d ";" | sed "s:|:\t:g" >> list_urban_nonsyn_genes.txt

grep HIGH urban_ann.vcf | cut -f8 | cut -f7 -d ";" | sed "s:|:\t:g" >> list_urban_nonsyn_genes.txt


sed -i -e "s:GL:chrUn_GL:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^c\t:chrLGc\t:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^a:chrLGa:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^b:chrLGb:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^d:chrLGd:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^f:chrLGf:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^g:chrLGg:" multinomial_SNPs_positions_outliers.txt
sed -i -e "s:^h:chrLGh:" multinomial_SNPs_positions_outliers.txt

plink --bfile filtered --pca --allow-extra-chr \
      --extract range urban_SNPs_positions_outliers.txt \
      --out urban_SNPs

plink --bfile filtered --pca --allow-extra-chr \
      --extract range multinomial_SNPs_positions_outliers.txt \
      --out multinomial_SNPs
```

In R, we can visualize the PCA on outlier SNPs.

```{r}
library(geodata)
library(terra)
library(adegenet)
library(vegan)
library(data.table)
library(ggplot2)
library(robust)
library(qvalue)
library(psych)
library(terra)
library(raster)
library(rworldmap)
library(dplyr)
library(caret)

coords = read.table("FEEMS.coord",h=F)
colnames(coords)=c("lon","lat")
pca=read.table("urban_SNPs.eigenvec")
eigenvals=read.table("urban_SNPs.eigenval")
(eigenvals$V1/sum(eigenvals$V1))[1:2]  
colnames(pca)=c("Individual","FID",paste("PC",seq(1:(ncol(pca)-2)),sep=""))
total=cbind(pca,coords)

colnumber=ncol(total)
map.polygon <- getMap(resolution = "low")


ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black") +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = PC1), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()




ggplot() +
    # Plot the map
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = PC1, y = PC2, fill = paste(lon,lat)), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
        coord_equal() + 
    xlab("PC1") +
    ylab("PC2") + 
    theme_bw()





coords = read.table("FEEMS.coord",h=F)
colnames(coords)=c("lon","lat")
pca=read.table("multinomial_SNPs.eigenvec")
eigenvals=read.table("multinomial_SNPs.eigenval")
(eigenvals$V1/sum(eigenvals$V1))[1:2]  
colnames(pca)=c("Individual","FID",paste("PC",seq(1:(ncol(pca)-2)),sep=""))
total=cbind(pca,coords)

colnumber=ncol(total)
map.polygon <- getMap(resolution = "low")


ggplot() +
    # Plot the map
    geom_path(data = map.polygon, aes(x = long, y = lat, group = group), color = "black") +
    
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = lon, y = lat, fill = PC1), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
    
    scale_fill_gradient2(low = "darkblue", high = "red") +  
    xlim(-91.5, -89.3) + 
    ylim(29, 31) +
    coord_equal() + 
    xlab("Longitude") +
    ylab("Latitude") + 
    theme_bw()


ggplot() +
    # Plot the map
    # Add lines from original position to jittered position
    # Plot jittered points
    geom_point(data = total, aes(x = PC1, y = PC2, fill = paste(lon,lat)), 
               size = 5, shape = 21, color = "black") +  # Black outline for visibility
        coord_equal() + 
    xlab("PC1") +
    ylab("PC2") + 
    theme_bw()



```

## Using WZA on RDA results to obtain estimates over 50kb non-overlapping windows.

First we need to prepare the input, which should be rather easy.

We need the allele frequencies for our markers:

```{bash}
plink --bfile filtered --allow-extra-chr --double-id  --freq --out filtered
```

"As input, the [general_WZA_script.py](https://github.com/TBooker/WZA/blob/master/general_WZA_script.py) requires one input file, a table of GEA scores for your SNPs. The script refers to this file as the"correlations" file.

This table is required to contain the following information:

-   Scores for individual SNPs (e.g. these could be *p*-values from a correlation analysis or Bayes factors)

-   A variable that indicates the windows that SNPs belong to (e.g. a column of gene names, or analysis window names)

-   The average minor allele frequency across populations"

We can generate the first two windows, using the Z scores for the first RD axis in the analysis correcting for population structure with 2 PCs.

```{r}
#RDA2=as.data.frame(fread("RDApartial_loadings_2PCs.txt.gz"))
RDA2=as.data.frame(fread("RDApartial_loadings.txt.gz"))

freq=as.data.frame(fread("filtered.frq"))
colnames(RDA2)=c("MRK","CHR","POSITION","score_RDA1","score_RDA2","score_RDA3","score_RDA4","CCA_RDA1","CCA_RDA2","CCA_RDA3","CCA_RDA4")

zscore <- function(x){
  score <- (x-mean(x))/ sd(x)     # find loadings +/-z sd from mean loading     
}

Z=cbind(RDA2[,2:3],zscore(RDA2$score_RDA1))
colnames(Z)=c("CHR","POSITION","Zscore")

chi_sq_values <- Z$Zscore^2
lambda <- median(chi_sq_values) / qchisq(0.5, df = 1)

Z$P_val =  2 * (1 - pnorm(abs(Z$Zscore)))/max(1,lambda)
options(scipen=999999)

Z$window=paste(Z$CHR,(floor(Z$POSITION/50e3)*50e3),sep="_")
Z$MAF=freq$MAF
colnames(Z)=c("CHR","POSITION","Zscore","pVal","window_names","MAF")
write.csv(Z[,4:6],file="P-values_WZA_2PCs_RDA1.csv",quote=F,col.names=T,row.names=F)



RDA=as.data.frame(fread("RDApartial_4RDAs.txt.gz"))

freq=as.data.frame(fread("filtered.frq"))
colnames(RDA)=c("CHR","POSITION","P_val","Q_val")
RDA$window=paste(RDA$CHR,(floor(RDA$POSITION/50e3)*50e3),sep="_")
Z=cbind(RDA[,c(1:3,5)],freq$MAF)
colnames(Z)=c("CHR","POSITION","pVal","window_names","MAF")

write.csv(Z,file="P-values_WZA_3PCs_RDA_Mahala_4RDAs.csv",quote=F,col.names=T,row.names=F)


```

Once done, on a PC:

```{bash}
          
conda activate pixy
 
cut -f3,18,19 LFMM_scan_K3_total.txt > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat P_val_combined \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_P_val_combined.csv \
                              --sep "\t" 
cut -f4,18,19 LFMM_scan_K3_total.txt > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat canopy_cover \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_canopy_cover.csv \
                              --sep "\t" 
cut -f6,18,19 LFMM_scan_K3_total.txt > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC1 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_env_PC1.csv \
                             --sep "\t" 

cut -f8,18,19 LFMM_scan_K3_total.txt > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC2 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_env_PC2.csv \
                              --sep "\t" 
cut -f10,18,19 LFMM_scan_K3_total.txt > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC3 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_env_PC3.csv \
                              --sep "\t" 
cut -f12,18,19 LFMM_scan_K3_total.txt > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat nightlight \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_nightlight.csv \
                              --sep "\t" 

cut -f14,18,19 LFMM_scan_K3_total.txt > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat per_impervious \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_per_impervious.csv \
                              --sep "\t" 
                              
cut -f16,18,19 LFMM_scan_K3_total.txt > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat urban_PC1 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 50kb_window_urban_PC1.csv \
                              --sep "\t" 
                              
                              
                              
###5kb windows
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f3,18,19 > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat P_val_combined \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_P_val_combined.csv \
                              --sep "\t" 
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 |cut -f4,18,19  > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat canopy_cover \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_canopy_cover.csv \
                              --sep "\t" 
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 |cut -f6,18,19 > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC1 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_env_PC1.csv \
                             --sep "\t" 

awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f8,18,19 > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC2 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_env_PC2.csv \
                              --sep "\t" 
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f10,18,19 > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat env_PC3 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_env_PC3.csv \
                              --sep "\t" 
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f12,18,19  > temp.txt

python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat nightlight \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_nightlight.csv \
                              --sep "\t" 

awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f14,18,19 > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat per_impervious \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_per_impervious.csv \
                              --sep "\t" 
                              
awk '{window=int($2/5000)*5000; print $0 "\t" $1 "_" window}' LFMM_scan_K3_total.txt | cut -f1-18,20 | cut -f16,18,19 > temp.txt
python ./WZA/general_WZA_script.py --correlations  temp.txt \
                              --summary_stat urban_PC1 \
                              --window CHR_0 \
                              --MAF MAF \
                              --output 5kb_window_urban_PC1.csv \
                              --sep "\t" 
                              
  
```

Then in R, for the Manhattan plots, and then GO term enrichment analysis.

```{r}
# Load required libraries
library(ggplot2)
library(ggrastr)
library(qvalue)
library(tidyr)
library(dplyr)
library(readr)
library(stringr)
library(data.table)
library(purrr)

# List all relevant CSV files

files <- list.files(pattern = "^50kb_window_.*\\.csv$")

gea_data <- map_dfr(files, function(file) {
  test_name <- str_extract(file, "(?<=50kb_window_).*(?=\\.csv)")

  read_csv(file, show_col_types = FALSE) %>%
    dplyr::select(gene, Z_pVal) %>%
    mutate(Test = test_name)
})


gea_data$gene <- gsub("chrLG", "", gea_data$gene)
gea_data$gene <- gsub("chrUn_", "", gea_data$gene)
gea_data$gene <- gsub("chr", "", gea_data$gene)


gea_data <- gea_data %>%
  separate(gene, into = c("CHR", "POSITION"), sep = "_") %>%
  mutate(POSITION = as.numeric(POSITION))


gea_data$CHR <- gsub("chrLG", "", gea_data$CHR)
gea_data$CHR <- gsub("chrUn_", "", gea_data$CHR)
gea_data$CHR <- gsub("chr", "", gea_data$CHR)

setDT(gea_data)

intervals_of_interest <- df_narrow
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)
# Use non-equi join
gea_data[, highlight := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight to "yes" for matching rows
gea_data[matches, highlight := "yes", on = .(CHR, POSITION)]

chr=c("4","2")
start=c(121495000,23410000)
end=c(123655000,25660000)
intervals_of_interest=as.data.frame(cbind(chr,start,end))

intervals_of_interest$chr=as.factor(intervals_of_interest$chr)
intervals_of_interest$start=as.integer(intervals_of_interest$start)
intervals_of_interest$end=as.integer(intervals_of_interest$end)


setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight2 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight2 to "yes" for matching rows
gea_data[matches, highlight2 := "yes", on = .(CHR, POSITION)]





intervals_of_interest <- df_extended
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight3 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight3 to "yes" for matching rows
gea_data[matches, highlight3 := "yes", on = .(CHR, POSITION)]



intervals_of_interest <- df_behav
colnames(intervals_of_interest)[1]="chr"

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight4 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight4 to "yes" for matching rows
gea_data[matches, highlight4 := "yes", on = .(CHR, POSITION)]




gea_data=subset(gea_data,(gea_data$CHR>0 & gea_data$CHR<7))


# Reshape gea_data to long format
desired_order <- c("P_val_combined", "env_PC1", "env_PC2", "env_PC3",
                   "urban_PC1", "canopy_cover", "nightlight", "per_impervious")

gea_data$Test <- factor(gea_data$Test, levels = desired_order)


pdf(file="GEA_Manhattan_pvalues_50kb.pdf",width=15,height=9)
ggplot(gea_data, aes(x = POSITION, y = -log10(Z_pVal), color = factor(CHR))) +
    geom_point_rast(size = 0.8) +
    scale_color_manual(values = rep(c("darkgrey", "black"), 16)) +
    geom_point_rast(data = subset(gea_data, highlight == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "darkgreen", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight2 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "red", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight3 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "darkorange", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight4 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "lightblue", size = 1.5) +
    facet_grid(Test ~ CHR, scales = "free_y") +
    theme_minimal() +
    labs(x = "Genomic Position", y = "-log10(p-value)", title = "Manhattan Plots by Test and Chromosome") +
    theme(legend.position = "none")
dev.off()



```

5kb windows

```{r}
# Load required libraries
library(ggplot2)
library(ggrastr)
library(qvalue)
library(tidyr)
library(dplyr)
library(readr)
library(stringr)
library(data.table)
library(purrr)

# List all relevant CSV files

files <- list.files(pattern = "^5kb_window_.*\\.csv$")

gea_data <- map_dfr(files, function(file) {
  test_name <- str_extract(file, "(?<=5kb_window_).*(?=\\.csv)")

  read_csv(file, show_col_types = FALSE) %>%
    dplyr::select(gene, Z_pVal) %>%
    mutate(Test = test_name)
})


gea_data$gene <- gsub("chrLG", "", gea_data$gene)
gea_data$gene <- gsub("chrUn_", "", gea_data$gene)
gea_data$gene <- gsub("chr", "", gea_data$gene)


gea_data <- gea_data %>%
  separate(gene, into = c("CHR", "POSITION"), sep = "_") %>%
  mutate(POSITION = as.numeric(POSITION))


gea_data$CHR <- gsub("chrLG", "", gea_data$CHR)
gea_data$CHR <- gsub("chrUn_", "", gea_data$CHR)
gea_data$CHR <- gsub("chr", "", gea_data$CHR)

setDT(gea_data)

intervals_of_interest <- df_narrow
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)
# Use non-equi join
gea_data[, highlight := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight to "yes" for matching rows
gea_data[matches, highlight := "yes", on = .(CHR, POSITION)]

chr=c("4","2")
start=c(121495000,23410000)
end=c(123655000,25660000)
intervals_of_interest=as.data.frame(cbind(chr,start,end))

intervals_of_interest$chr=as.factor(intervals_of_interest$chr)
intervals_of_interest$start=as.integer(intervals_of_interest$start)
intervals_of_interest$end=as.integer(intervals_of_interest$end)


setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight2 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight2 to "yes" for matching rows
gea_data[matches, highlight2 := "yes", on = .(CHR, POSITION)]





intervals_of_interest <- df_extended
colnames(intervals_of_interest)[1]="chr"
# Mark the intervals of interest

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight3 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight3 to "yes" for matching rows
gea_data[matches, highlight3 := "yes", on = .(CHR, POSITION)]



intervals_of_interest <- df_behav
colnames(intervals_of_interest)[1]="chr"

setDT(intervals_of_interest)

# Add key columns
setkey(intervals_of_interest, chr, start, end)
setkey(gea_data, CHR, POSITION, POSITION)

# Use non-equi join
gea_data[, highlight4 := "no"]
matches <- foverlaps(gea_data[, .(CHR, POSITION, POS_END = POSITION)], 
                     intervals_of_interest[, .(chr, start, end)],
                     by.x = c("CHR", "POSITION", "POS_END"),
                     by.y = c("chr", "start", "end"),
                     nomatch = 0L)

# Set highlight4 to "yes" for matching rows
gea_data[matches, highlight4 := "yes", on = .(CHR, POSITION)]




gea_data=subset(gea_data,(gea_data$CHR>0 & gea_data$CHR<7))


# Reshape gea_data to long format
desired_order <- c("P_val_combined", "env_PC1", "env_PC2", "env_PC3",
                   "urban_PC1", "canopy_cover", "nightlight", "per_impervious")

gea_data$Test <- factor(gea_data$Test, levels = desired_order)


pdf(file="GEA_Manhattan_pvalues_5kb.pdf",width=15,height=9)
ggplot(gea_data, aes(x = POSITION, y = -log10(Z_pVal), color = factor(CHR))) +
    geom_point_rast(size = 0.8) +
    scale_color_manual(values = rep(c("darkgrey", "black"), 16)) +
    geom_point_rast(data = subset(gea_data, highlight == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "darkgreen", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight2 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "red", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight3 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "darkorange", size = 1.5) +
    geom_point_rast(data = subset(gea_data, highlight4 == "yes"),
                    aes(x = POSITION, y = -log10(Z_pVal)), color = "lightblue", size = 1.5) +
    facet_grid(Test ~ CHR, scales = "free_y") +
    theme_minimal() +
    labs(x = "Genomic Position", y = "-log10(p-value)", title = "Manhattan Plots by Test and Chromosome") +
    theme(legend.position = "none")
dev.off()



```
